Sequence memory is an information constraint: before a predictor can use a distinction, its runtime state must make that distinction observable. This paper develops Concept Graph Attention (CGA) from that first-principles view. We formalize a runtime state as an information partition, separate \emph{repair} updates that change payloads inside a fixed partition from \emph{refinement} updates that change the partition itself, and prove a repair-only obstruction: if the current partition merges cases whose refined Bayes actions differ, no amount of local repair, caching, or optimizer sophistication can remove the resulting linear regret against the refined oracle.

The theory identifies which distinctions are worth adding. A candidate bit has value only through risk reduction, not through geometric novelty or frequency. Under log loss, the population value of a binary distinction is $I(Y;U\mid\Pi)$; under squared loss, it is the conditional mean-separation energy induced by the split. We then give a guarded empirical refinement rule that accepts structural growth only when a held-out, cost-penalized refinement gain beats the best repair gain by a finite-sample margin. The full appendix gives the measure-theoretic join form, finite-class oracle inequalities, graph data-structure bounds, and projection-memory/CPM bridge proofs.

CGA is the sequence-modeling specialization of this refine-or-repair program. It keeps attention's continuous readout but changes the memory basis from prefix tokens to a bounded bank of concept identities. Each concept has a fixed address and a runtime value memory. Tokens read from the concept bank with attention-style softmax weights and write back with gated continuous updates and trainable retention. During training, gradients flow through the read, write, gate, and retention parameters; during inference, model weights stay fixed while the concept value memory is updated by the same forward rule. Empirically, we evaluate CGA through state-equalized MQAR and matched language-model pretraining protocols against modern efficient sequence mixers.
