\section{Mathematical Model: Partitions, Predictors, and Bayes Risk}
\label{sec:preliminaries}

We first formalize the information available to a runtime learner as a partition of the case space. The measurable-space notation used later is the direct sigma-algebra generalization, but the computational object is a finite partition.

\begin{table}[t]
\centering
\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.20\linewidth}p{0.70\linewidth}}
\toprule
Symbol & Meaning \\
\midrule
$\Omega$ & finite case space \\
$P$ & distribution over cases \\
$Y:\Omega\to\calY$ & consequence or label variable \\
$\mathsf A$ & action space \\
$L:\calY\times\mathsf A\to\R$ & loss \\
$\Pi$ & finite information partition of $\Omega$ \\
$\Pred(\Pi)$ & predictors measurable with respect to $\Pi$ \\
$R_L(f)$ & risk of predictor $f$ \\
$R_L(\Pi)$ & Bayes risk achievable under partition $\Pi$ \\
$I:\Omega\to\calZ$ & information map inducing a partition $\Pi(I)$ \\
$U:\Omega\to\{0,1\}$ & binary distinction \\
$\Pi\vee U$ & partition obtained by splitting $\Pi$ by $U$ \\
$\calF,\calG$ & sigma-algebras or information structures \\
$V_L(\calF,\calG)$ & value of refining $\calF$ to $\calG$ \\
$d_L^\epsilon$ & loss-dependent distinction complexity \\
$a_j,c_j,M_j$ & DGM anchor, centroid, and local memory for concept $j$ \\
$E_t,g_e,\calP_t^G$ & DGM edge set, edge generator, and graph-generated partition \\
$\calB_t^{\mathrm{score}}$ & held-out scoring buffer for empirical refinement tests \\
$\widehat\Delta_{\mathrm{repair}},\widehat\Delta_{\mathrm{refine}}$ & empirical repair and refinement gains \\
\bottomrule
\end{tabular}
\caption{Notation used throughout the paper. We say $\Pi'$ is finer than $\Pi$ if every atom of $\Pi'$ is contained in an atom of $\Pi$.}
\label{tab:notation}
\end{table}

\subsection{Partitions, predictors, and risk}

Let $\Omega$ be a finite case space, let $P$ be a distribution on $\Omega$, let $Y:\Omega\to\calY$ be the consequence variable, let $\mathsf A$ be an action space, and let $L:\calY\times\mathsf A\to\R$ be a loss with finite expectations for the predictors under discussion.

\begin{definition}[Information partition and partition-measurable predictor]
\label{def:partition-measurable}
A finite information partition is a collection of nonempty disjoint atoms
\begin{align*}
\Pi=\{B_1,\ldots,B_m\}
\end{align*}
whose union is $\Omega$. It induces the equivalence relation $\omega\sim_\Pi\omega'$ iff $\omega,\omega'$ lie in the same atom $B\in\Pi$.

A predictor is $\Pi$-measurable if it cannot distinguish states inside the same atom:
\begin{align*}
\Pred(\Pi)
=
\{f:\Omega\to\mathsf A:
\omega\sim_\Pi\omega'\Rightarrow f(\omega)=f(\omega')\}.
\end{align*}
Equivalently, a $\Pi$-measurable predictor is a map $a:\Pi\to\mathsf A$ evaluated on the atom containing $\omega$.
\end{definition}

\begin{definition}[Bayes risk under a partition]
\label{def:partition-bayes-risk}
For $f:\Omega\to\mathsf A$, write $R_L(f)=\E[L(Y,f)]$. The Bayes risk achievable under partition $\Pi$ is
\begin{align}
R_L(\Pi)
=
\inf_{f\in\Pred(\Pi)}
\E[L(Y,f)]
=
\sum_{B\in\Pi}P(B)
\inf_{a\in\mathsf A}\E[L(Y,a)\mid \omega\in B],
\label{eq:partition-bayes-risk}
\end{align}
where atoms with $P(B)=0$ may be assigned arbitrary actions. Thus $R_L(\Pi)$ is the best possible risk after the representation has committed to the equivalence relation $\sim_\Pi$.
\end{definition}

\begin{definition}[Information map and induced partition]
\label{def:information-map-partition}
An information map is a runtime representation $I:\Omega\to\calZ$. It induces the partition
\begin{align*}
\Pi(I)=\{I^{-1}(z):z\in\calZ,\ I^{-1}(z)\ne\varnothing\}.
\end{align*}
Any predictor that factors through $I$, written $f(\omega)=a(I(\omega))$, is measurable with respect to $\Pi(I)$.
\end{definition}

\section{Online Prediction with Repair and Refinement}
\label{sec:online-model}

The online learner has two update types: repair preserves the current partition, while refinement joins it with a new binary distinction.

\begin{definition}[Binary distinction and partition join]
\label{def:partition-join}
A binary distinction is a function $U:\Omega\to\{0,1\}$. Joining $\Pi$ with $U$ means splitting every atom of $\Pi$ according to the value of $U$:
\begin{align}
\Pi\vee U
=
\{B\cap U^{-1}(0),B\cap U^{-1}(1):B\in\Pi\}\setminus\{\varnothing\}.
\label{eq:partition-join}
\end{align}
\end{definition}

\begin{definition}[Refinable online prediction problem]
\label{def:refinable-online-prediction}
A refinable online prediction problem is a tuple
\begin{align*}
\mathfrak P=(\Omega,\calY,\mathsf A,P,L,\calO,\calU,c),
\end{align*}
where $\Omega$ is finite, $Y:\Omega\to\calY$ is the consequence variable, $P$ is the sampling distribution, $L:\calY\times\mathsf A\to\R$ has finite risk for admissible predictors, $\calO\subseteq2^\Omega$ is the pre-outcome observable sigma-algebra, $\calU\subseteq\{0,1\}^{\Omega}$ is a class of admissible binary distinctions that are $\calO$-measurable, and $c:\calU\to\R_+$ assigns refinement costs. Bounded losses are required only in finite-sample concentration statements, where they are stated explicitly. The $\calO$-measurability condition prevents a prediction-time distinction from depending directly on the consequence label; pure value-of-information examples can be recovered by taking $\calO=2^\Omega$.

At round $t$, the learner maintains $S_t=(\Pi_t,a_t)$ with $a_t:\Pi_t\to\mathsf A$. A case $\omega_t\sim P$ is drawn, the learner observes only the atom $B_t\in\Pi_t$ containing $\omega_t$, predicts $a_t(B_t)$, observes $Y_t=Y(\omega_t)$, and then chooses one of two updates:
\begin{align*}
\mathrm{Repair}:&\qquad \Pi_{t+1}=\Pi_t,\quad a_{t+1}\text{ may change},\\
\mathrm{Refine}:&\qquad \Pi_{t+1}=\Pi_t\vee U_t\text{ for some }U_t\in\calU.
\end{align*}
The cumulative objective is prediction loss plus refinement cost,
\begin{align*}
\calL_T
=
\sum_{t=1}^T L(Y_t,a_t(B_t))
+
\lambda\sum_{t:\mathrm{Refine}}c(U_t).
\end{align*}
\end{definition}

\begin{definition}[Fixed-information repair update]
\label{def:fixed-information-repair}
Let a parameterized runtime predictor factor through a fixed information map $I:\Omega\to\calZ$:
\begin{align*}
f_\psi(\omega)=a_\psi(I(\omega)).
\end{align*}
An update of $\psi$ is a fixed-information repair update if $I$ is unchanged. The update rule may be a gradient step, a closed-form sufficient-statistic update, a cache write, or any other optimizer; what makes it repair is the invariance of the induced partition $\Pi(I)$, not the optimizer used.
\end{definition}

\section{The Repair-Only Obstruction}
\label{sec:repair-obstruction}

If the partition is fixed, the learner cannot beat the Bayes risk of that partition, no matter how sophisticated the repair update is.

\begin{definition}[Repair gap and refinement gap]
\label{def:repair-refinement-gaps}
For a concrete predictor $f\in\Pred(\Pi)$ and a candidate refined partition $\Pi^\star$ that refines $\Pi$, define
\begin{align}
\Delta_{\mathrm{repair}}(f,\Pi)
&=R_L(f)-R_L(\Pi),
\label{eq:repair-gap}\\
\Delta_{\mathrm{refine}}(\Pi,\Pi^\star)
&=R_L(\Pi)-R_L(\Pi^\star).
\label{eq:refinement-gap}
\end{align}
The first gap is the residual error of the current predictor at fixed equivalence relation. The second is the value of changing the equivalence relation.
\end{definition}

\begin{theorem}[Runtime error decomposition]
\label{thm:error-decomposition}
Let $f_t\in\Pred(\Pi_t)$ and let $\Pi^\star$ refine $\Pi_t$. Then
\begin{align}
R_L(f_t)-R_L(\Pi^\star)
=
\Delta_{\mathrm{repair}}(f_t,\Pi_t)
+
\Delta_{\mathrm{refine}}(\Pi_t,\Pi^\star).
\label{eq:error-decomposition}
\end{align}
Both terms are nonnegative.
\end{theorem}

\begin{proof}
Add and subtract $R_L(\Pi_t)$. The repair gap is nonnegative because $f_t$ is feasible for the infimum defining $R_L(\Pi_t)$. The refinement gap is nonnegative because every $\Pi_t$-measurable predictor is also measurable with respect to any refinement $\Pi^\star$.
\end{proof}

\subsection{Repair-only lower bound}

The pointwise contradiction inside a cell becomes a sequential lower bound. If the current partition collapses states that a refined oracle separates, no repair-only learner can avoid paying the corresponding Bayes-risk gap every round.

\begin{theorem}[Repair-only linear regret]
\label{thm:repair-only-regret}
Let $\Pi$ be a fixed partition and let $\Pi^\star$ be a refinement. Consider any learner whose round-$t$ prediction $\hat f_t$ is always in $\Pred(\Pi)$, possibly chosen adaptively from all previous observations. Then, for iid $\omega_t\sim P$,
\begin{align*}
\E\left[\sum_{t=1}^T L(Y_t,\hat f_t(\omega_t))\right]
\ge
T R_L(\Pi).
\end{align*}
Consequently, if Bayes actions exist under $\Pi^\star$, then relative to an oracle using $\Pi^\star$,
\begin{align*}
\E[\mathrm{Regret}_T]
\ge
T\left(R_L(\Pi)-R_L(\Pi^\star)\right).
\end{align*}
If the infimum under $\Pi^\star$ is not attained, the same statement holds in the epsilon-oracle sense: for every $\epsilon>0$, regret against an $\epsilon$-optimal $\Pi^\star$-measurable oracle is at least $T(R_L(\Pi)-R_L(\Pi^\star)-\epsilon)$. Thus if $R_L(\Pi)>R_L(\Pi^\star)$, every repair-only learner suffers $\Omega(T)$ expected regret against refined or epsilon-refined oracles.
\end{theorem}

\begin{proof}
Condition on the learner's history before round $t$. The chosen predictor $\hat f_t$ is $\Pi$-measurable, so by definition of $R_L(\Pi)$,
\begin{align*}
\E[L(Y_t,\hat f_t(\omega_t))\mid \text{history}]
\ge
R_L(\Pi).
\end{align*}
Taking expectation and summing over $t$ gives the first inequality. If Bayes actions exist on the atoms of $\Pi^\star$, the $\Pi^\star$ oracle achieves $R_L(\Pi^\star)$ per round by using them, so subtracting its cumulative risk gives the regret bound. Without attained Bayes actions, for every $\epsilon>0$ choose an $\epsilon$-optimal atomwise predictor under $\Pi^\star$; subtracting its cumulative risk gives the epsilon-oracle bound.
\end{proof}

The displayed comparator gives the refined partition to the oracle for free. If regret includes a one-time refinement cost $C(\Pi^\star)$ for reaching $\Pi^\star$, the same argument gives
\begin{align*}
\E[\mathrm{Regret}_T]
\ge
T\left(R_L(\Pi)-R_L(\Pi^\star)\right)-\lambda C(\Pi^\star).
\end{align*}

\begin{corollary}[Fixed-information gradient repair lower bound]
\label{cor:fixed-information-gradient-repair}
Consider any online optimizer, including backpropagation, whose predictions at every round factor through a fixed information map $I$:
\begin{align*}
\hat f_t(\omega)=a_{\psi_t}(I(\omega)).
\end{align*}
Then $\hat f_t\in\Pred(\Pi(I))$ for all $t$, so Theorem~\ref{thm:repair-only-regret} applies with $\Pi=\Pi(I)$. In particular, if a refined partition $\Pi^\star$ has $R_L(\Pi(I))>R_L(\Pi^\star)$, this fixed-information gradient repair suffers linear regret against the refined oracle. Backpropagation through parameters that change $I$ is outside this corollary; in the present model, such an update changes the information structure rather than merely repairing actions on fixed atoms.
\end{corollary}

\begin{example}[Two-state separation]
\label{ex:two-state-separation}
Let $\Omega=\{0,1\}$, $P(0)=P(1)=1/2$, and $Y=\omega$. Under the trivial partition $\Pi=\{\{0,1\}\}$, the best log-loss predictor is $q(Y=0)=q(Y=1)=1/2$, so $R_{\log}(\Pi)=\log 2$. The refined partition $\Pi^\star=\{\{0\},\{1\}\}$ has $R_{\log}(\Pi^\star)=0$ in the noiseless case when point-mass predictions are allowed. Therefore any repair-only learner has expected regret $T\log 2$ against the refined oracle. This example is not about neural networks, caches, or trees. It is about equivalence relations: a learner that identifies two states cannot assign them two different consequences.
\end{example}

\begin{corollary}[Contradiction requires distinction]
\label{thm:contradiction:informal}
Let $\Pi$ be a partition and let $y:\Omega\to\mathsf A$ be a required deterministic output. If $\omega,\omega'$ lie in the same atom of $\Pi$ but $y(\omega)\ne y(\omega')$, then no predictor in $\Pred(\Pi)$ satisfies both requirements.
\end{corollary}

\begin{proof}
Every $f\in\Pred(\Pi)$ is constant on the atom containing $\omega$ and $\omega'$, so $f(\omega)=f(\omega')$. Satisfying both requirements would imply $y(\omega)=y(\omega')$, a contradiction.
\end{proof}

\subsection{Sigma-algebra extension}

The finite partition model is the main computational model. The sigma-algebra notation is the measure-theoretic extension used for value identities and language-model interpretations. In this formulation, $L$ is measurable with finite expectation for the actions considered.

\begin{definition}[$\calF$-measurable predictor and Bayes action]
\label{def:F-measurable-predictor}
On a probability space $(\Omega,\calE,\mathbb P)$, an information structure is a sub-sigma-algebra $\calF\subseteq\calE$. For a measurable action space $(\mathsf A,\Sigma_{\mathsf A})$, a predictor $a:\Omega\to\mathsf A$ is $\calF$-measurable if
\begin{align*}
a^{-1}(C)\in\calF
\qquad
\text{for every }C\in\Sigma_{\mathsf A}.
\end{align*}
The admissible predictor class and Bayes risk are
\begin{align*}
\Pred(\calF)
=
\{a:\Omega\to\mathsf A:a\text{ is }(\calF,\Sigma_{\mathsf A})\text{-measurable}\},
\qquad
R_L(\calF)=\inf_{a\in\Pred(\calF)}\E[L(Y,a)].
\end{align*}
When the infimum is attained, any minimizer is an $\calF$-Bayes action and is denoted $a_\calF^\star$. A random variable $b$ has an $\calF$-measurable version if there exists an $\calF$-measurable random variable $\tilde b$ such that $\tilde b=b$ almost surely.
\end{definition}

For a finite partition $\Pi$, Definition~\ref{def:F-measurable-predictor} recovers Definition~\ref{def:partition-measurable} by taking $\calF=\sigma(\Pi)$ and the action space discrete.

Implementations often have fixed pre-existing prediction outputs that are not created by the runtime refinement mechanism, such as a backbone distribution or other fixed pre-outcome features used outside DGM routing. We write this base information as a sigma-algebra $\calB$ and measure runtime refinements through the total information $\calB\vee\calF_t$. The full hidden state used to propose or instantiate distinctions should not automatically be included as freely conditionable base information: if $\calB$ contains $h$, then every edge predicate $g_e(h)$ and anchor-order event is already $\calB$-measurable, so the unrestricted sigma-algebra refinement gap becomes trivial. Thus the refinement gap relative to a target $\calF^\star$ is
\begin{align*}
R_L(\calB\vee\calF_t)-R_L(\calB\vee\calF^\star).
\end{align*}
Repairs keep $\calF_t$ fixed, while refinements replace $\calF_t$ by $\calF_t\vee\sigma(U_t)$. All monotonicity and value identities below apply after joining both sides with the same fixed $\calB$.

Section~\ref{sec:value} uses this notation to characterize exactly when a richer information structure is useful.
