\section{Conceptual Attention as a Data Structure for Refinable Partitions}
\label{sec:algorithms}

This appendix describes the graph-growing CA extension that realizes literal partition refinement. The main experiments use the simpler fixed-bank specialization from Section~\ref{sec:main-ca}: concept identities are preallocated, all slots are active, edge gates are disabled, and the online operation is gated value-memory repair inside a bounded concept bank. The distinction matters for claims. The fixed-bank layer tests whether preallocated concept capacity and forward-time repair can replace token-indexed memory; the graph-growing extension below is the implementation route for accepting new distinctions online.

In the graph-growing extension, CA implements refinable information structures as a graph: edges generate distinctions and nodes store repairable memories. The ingredients are related to decision-tree refinement, adaptive category creation, prototype memories, and external neural memories, but CA uses them to implement the repair/refinement distinction \citep{breiman1984classification,domingos2000mining,carpenter1987massively,grossberg2013adaptive,kohonen1990self,weston2015memory,graves2016hybrid}. The theory so far gives a diagnostic: decide whether an observed error is better explained by $\Delta_{\mathrm{repair}}$ at fixed information or by $\Delta_{\mathrm{refine}}$ from missing distinctions. A runtime implementation therefore needs a state representation that can both repair local statistics and add distinctions. Graph-growing CA is the data-structure realization of the abstract refinable-partition model.

\subsection{Fixed-bank specialization and block kernels}
\label{sec:fixed-bank-kernel}

The fixed-bank layer used for MQAR and CALM can be read as a finite concept-memory state machine. For a block of length $B$, heads $H$, concepts $C$, address width $d_a$, and value width $d_v^c$, the dense read computes
\begin{align*}
S_{b,thj}=\gamma q_{b,th}^{\top}a_{hj},
\qquad
\alpha_{b,thj}=\operatorname{softmax}_j S_{b,thj},
\qquad
z_{b,th}=\sum_{j=1}^{C}\alpha_{b,thj}M_{b,hj}.
\end{align*}
The dense write computes delayed write scores
\begin{align*}
R_{b,thj}=\gamma \bar k_{b,th}^{\top}a_{hj},
\qquad
\beta_{b,thj}=\operatorname{softmax}_j R_{b,thj},
\end{align*}
and updates
\begin{align*}
M_{b+1,hj}
=
\rho_{bh}M_{b,hj}
+
\sum_{t=1}^{B}m_{bt}g_{bt}\beta_{b,thj}v_{b,th},
\end{align*}
where $m_{bt}$ masks invalid delayed writes. The online state is only $M_b$, with size $C H d_v^c$; the concept identities $a_{hj}$ are fixed buffers or parameters of the layer, not sequence state.

The arithmetic cost per block is
\begin{align*}
O(B H C d_a + B H C d_v^c)
\end{align*}
for the read and the same order for the dense write, plus projection costs shared with the surrounding decoder block. The Triton read kernel used in the current code tiles the concept, address, value, and block dimensions and applies the same equations under the implementation bounds used by the CUDA path, namely block length at most $128$, concept count at most $256$, address width at most $128$, and value width at most $256$. These bounds are engineering constraints of the fused dense kernel, not assumptions in the partition theory. Proposition~\ref{prop:main-block-ca} gives the mathematical interface: block size one is token-streaming CA, while larger blocks trade within-block memory visibility for parallelism.

\subsection{Why a graph, not a tree?}

\begin{definition}[CA concept]
\label{def:dgm-concept}
A CA concept is a persistent address for a prediction-relevant information cell, equipped with a local repairable memory. In the graph-growing formalization it is selected by a graph-generated hard-route map. It is not claimed to be a human-interpretable semantic concept and need not correspond one-to-one with an atom of the edge partition. Local memories live at graph nodes, and distinction predicates live on edges between nodes that have been confused.
\end{definition}

The central object is
\begin{align*}
\text{CA}=\text{graph-generated information structure}+\text{local memory repair}.
\end{align*}
The graph is not metadata: its edges generate a finite edge-skeleton information structure and gate the read path, while anchor order completes the exact hard route. The partition and generated-sigma-algebra language follows standard finite measurable-space constructions \citep{billingsley1995probability,kallenberg2002foundations}. The main-text statements below give the operational theorem forms; full proofs are in Appendix~\ref{app:algorithm-bridge-proofs}.

\begin{definition}[Hard-CA and Soft-CA]
\label{def:hard-soft-dgm}
Hard-CA is the formal algorithm: fixed anchors and accepted binary edges generate an edge skeleton, while the exact hard route is measured by the join of this edge skeleton with the anchor-distance ordering. A repair changes only local actions or memories and leaves the hard-route information fixed; a refinement adds an anchored concept, new edge generators, and the anchor-order events involving the new anchor. Soft-CA is the practical relaxation used for sparse neural hidden states: movable centroids retrieve candidates, smooth edge gates approximate hard eligibility, and routing weights mix local memories.
\end{definition}

The partition-level guarantees are exact for Hard-CA. Soft-CA is covered only as an approximation under the candidate-recall and margin conditions in Theorem~\ref{thm:soft-routing-hard-limit:informal}.

\subsection{Hard-CA state and invariants}

\begin{definition}[Hard-CA state and anchor edge]
\label{def:hard-dgm-state}
Let $\phi_\theta:\calX\to\R^d$ be an encoder, write $h=\phi_\theta(x)$, and let $\mathsf H\subseteq\R^d$ denote the hidden-state space on which the runtime structure is evaluated. A Conceptual Attention state is
\begin{align*}
\calS_t=\left(\{a_j,c_j,M_j,n_j\}_{j=1}^{N_t},G_t\right),
\qquad
G_t=(V_t,E_t),\qquad V_t=\{1,\ldots,N_t\}.
\end{align*}
Each component has a role forced by the refine-or-repair split. A runtime concept needs a stable identity, so $a_j\in\R^d$ is a frozen distinction anchor. Retrieval may adapt as examples arrive, so $c_j\in\R^d$ is a movable centroid used only by the soft implementation. State inside an adequate concept must be repairable, so $M_j$ is a local memory and $n_j$ is a usage count. If a concept is created because it was confused with existing concepts, that conflict becomes an edge in $E_t$. A new concept is initialized with $a_j=c_j=h$. Subsequent repair may move $c_j$, but anchors remain fixed once created, so the hard information structure is unchanged by centroid repair.

An edge $e=(i,j)$ stores a linear distinction
\begin{align*}
u_e=\frac{a_i-a_j}{\norm{a_i-a_j}_2},\qquad b_e=\frac{1}{2}u_e^\top(a_i+a_j),
\end{align*}
with side labels $s_{e,i}=+1$ and $s_{e,j}=-1$. Its binary generator is
\begin{align*}
g_e(h)=\one\{u_e^\top h>b_e\}.
\end{align*}
\end{definition}

\begin{definition}[Graph signature and edge-generated partition]
\label{def:graph-signature-partition}
The edge set induces the \emph{edge skeleton} information structure on $\mathsf H$. The graph signature is
\begin{align*}
\sigma_t^G(h)=\big(g_e(h)\big)_{e\in E_t}\in\{0,1\}^{E_t}.
\end{align*}
Two hidden states are indistinguishable by the current graph if their signatures agree. The nonempty level sets
\begin{align*}
A_a^t=\{h\in\mathsf H:\sigma_t^G(h)=a\},\qquad a\in\{0,1\}^{E_t},
\end{align*}
form a finite partition $\calP_t^G$ of hidden-state space, and the generated sigma-algebra is $\sigma(\calP_t^G)$. Adding a new edge is therefore a join with one binary generator:
\begin{align*}
\calP_{t+1}^G=\calP_t^G\vee\sigma(g_e).
\end{align*}
\end{definition}

This edge partition is only the skeleton. The exact hard route also needs the ordering of eligible anchors by anchor distance; that order is added below as $\calF_t^{\mathrm{anchor\text{-}order}}$. Centroids and local memories are adaptive statistics attached to hard-route states or sparse neighborhoods, not generators of the exact Hard-CA information structure. Concept nodes and graph-generated partition atoms are related but not identical: a node is a data-structure representative with memory and retrieval statistics; the formal information structure is generated by edge signatures and anchor-order events.

\begin{figure}[t]
\centering
\begin{tikzpicture}[x=1cm,y=1cm,>=stealth]
\tikzset{
    anchorpt/.style={circle,fill=black,inner sep=0pt,minimum size=4.2pt},
    centroid/.style={circle,draw=blue!75,thick,inner sep=0pt,minimum size=9pt},
    memory/.style={draw,rounded corners=2pt,fill=green!10,minimum width=0.85cm,minimum height=0.38cm,font=\scriptsize},
    lab/.style={font=\scriptsize},
    query/.style={circle,fill=orange!85,inner sep=0pt,minimum size=5pt}
}

\node[anchorpt,label={[lab]225:$a_i$}] (ai) at (0.9,0.9) {};
\node[anchorpt,label={[lab]20:$a_j$}] (aj) at (4.0,1.25) {};
\node[anchorpt,label={[lab]80:$a_k$}] (ak) at (2.55,3.0) {};
\draw[thick] (ai) -- (aj);
\draw[thick] (aj) -- (ak);

\draw[dashed,gray!70] (2.35,0.25) -- (2.12,2.25);
\draw[dashed,gray!70] (3.1,2.02) -- (4.72,3.35);

\node[centroid,label={[lab,blue!75]35:$c_i$}] (ci) at (1.32,1.22) {};
\node[centroid,label={[lab,blue!75]260:$c_j$}] (cj) at (3.72,0.94) {};
\node[centroid,label={[lab,blue!75]200:$c_k$}] (ck) at (2.18,2.68) {};

\node[memory] (mi) at (0.72,1.78) {$M_i$};
\node[memory] (mj) at (4.88,0.65) {$M_j$};
\node[memory] (mk) at (1.78,3.55) {$M_k$};

\node[query,label={[lab]0:query}] (q) at (5.35,2.7) {};
\draw[->,orange!85,thick] (q) -- (ak);
\draw[->,orange!85,thick] (q) -- (aj);

\node[font=\scriptsize,align=center,text width=6.0cm] at (2.85,0.0)
{distinction edges define the information structure; centroids retrieve; memories repair};
\end{tikzpicture}
\caption{CA state. Anchors are stable concept identities, centroids are movable retrieval statistics, distinction edges generate the information structure, and concept memories are repaired after a concept is judged adequate.}
\label{fig:dgm-state}
\end{figure}

\begin{theorem}[Anchor distinction separates its endpoints, informal version of Theorem~\ref{thm:anchor-edge:formal}]
\label{thm:anchor-edge:informal}
If $a_i\ne a_j$ and $u_e,b_e$ are defined as above for edge $e=(i,j)$, then
\begin{align*}
u_e^\top a_i-b_e=\frac{1}{2}\norm{a_i-a_j}_2,
\qquad
u_e^\top a_j-b_e=-\frac{1}{2}\norm{a_i-a_j}_2.
\end{align*}
Thus each edge distinction separates the two anchors that caused it.
\end{theorem}

Implementations should create such an edge only when the anchor separation clears a numerical guard,
\begin{align*}
\norm{a_i-a_j}_2>\eta_{\mathrm{anchor}},
\end{align*}
for a fixed $\eta_{\mathrm{anchor}}>0$. If a proposed new anchor is duplicate or nearly duplicate with a confused neighbor, the algorithm should skip that edge and fall back to repair, merge, or another admissible distinction.

\begin{theorem}[CA graph generates a finite information structure, informal version of Theorem~\ref{thm:dgm-graph-partition:formal}]
\label{thm:dgm-graph-partition:informal}
For any finite edge set $E_t$, equality of signatures $\sigma_t^G(h)$ is an equivalence relation whose classes form a finite edge-skeleton partition $\calP_t^G$. Adding one edge $g$ refines this skeleton by $\calP_t^G\vee\sigma(g)$.
\end{theorem}

\subsection{Soft-CA edge-gated prediction}

The hard route selects the nearest eligible anchor using fixed anchors, accepted edges, and anchor-distance order, so repairs that change only memories and centroids do not change $\calF_t^{\mathrm{CA}}$. Given $h$, Soft-CA instead computes retrieval distances $d_j(h)=\norm{h-c_j}_2^2$ to movable centroids. Moving centroids can change the smooth routing weights and therefore the actual Soft-CA predictor, even when the Hard-CA skeleton is fixed. Thus centroid repair is a repair in the formal hard-route structure, not a strict claim that the soft read map remains partition-measurable. A centroid-only top-$k$ set is not sufficient for the formal hard-route guarantee: if the hard eligible anchor never enters the candidate set, edge gates cannot recover it. Approximate-nearest-neighbor systems can make the retrieval term sublinear or hardware-efficient, but the formal guarantee still needs recall of the hard-route candidate \citep{indyk1998approximate,johnson2019billion}. We therefore use a recall-oriented candidate set
\begin{align*}
N(h)=
N_{k_c}^{\mathrm{centroid}}(h)
\cup
N_{k_a}^{\mathrm{anchor}}(h)
\cup
\operatorname{GraphNeighbors}\left(N_{k_c}^{\mathrm{centroid}}(h)\right),
\end{align*}
with duplicates removed and an optional budget cap after the union. The first term gives adaptive locality, the second protects the anchor-defined hard route, and the third recovers concepts connected by recent consequence conflicts. In exact small-scale implementations one may take $N(h)=\{1,\ldots,N_t\}$. If $\operatorname{Inc}_t(j)$ is the set of edges incident to concept $j$, define the soft compatibility of $h$ with concept $j$ by
\begin{align*}
\log \chi_j(h)=
\frac{1}{\sqrt{\max\{1,\abs{\operatorname{Inc}_t(j)}\}}}
\sum_{e\in \operatorname{Inc}_t(j)}
\beta_e\log \operatorname{sigmoid}\left(\frac{s_{e,j}(u_e^\top h-b_e)}{\tau_e}\right),
\end{align*}
where structural distinction edges have $\beta_e\in(0,1]$ and the empty sum gives $\chi_j(h)=1$. The degree normalization avoids systematically suppressing high-degree concepts while preserving hard rejection in the low-temperature limit. The edge-gated routing weights are
\begin{align}
\alpha_j(h)=
\frac{\exp(-d_j(h)/T)\chi_j(h)}
{\sum_{r\in N(h)}\exp(-d_r(h)/T)\chi_r(h)},
\qquad j\in N(h).
\label{eq:edge-gated-routing}
\end{align}
The prediction is the local mixture
\begin{align*}
\hat y=\sum_{j\in N(h)}\alpha_j(h)M_j(h).
\end{align*}

The hard route makes the information-structure connection explicit. A concept $j$ is eligible for $h$ if every incident edge bit places $h$ on the side recorded for $j$:
\begin{align*}
\begin{cases}
g_e(h)=1,& s_{e,j}=+1,\\
g_e(h)=0,& s_{e,j}=-1,
\end{cases}
\qquad\text{for every }e\in \operatorname{Inc}_t(j).
\end{align*}
This convention assigns boundary points to the nonpositive side because $g_e(h)=\one\{u_e^\top h>b_e\}$. The exact hard route used in the refinement theory routes to the nearest eligible anchor, falling back to the nearest anchor if no eligible concept is available. The soft read in \eqref{eq:edge-gated-routing} is the trainable implementation: it may retrieve by movable centroids, but those centroids are retrieval statistics rather than formal information cells.

The distinction partition alone does not determine the whole exact hard route, because eligible concepts still need to be ordered. Anchor distances provide this stable order geometry. Let $\calF_t^{\mathrm{edge}}=\sigma(g_e:e\in E_t)$ and let $\calF_t^{\mathrm{anchor\text{-}order}}$ be the finite sigma-algebra generated by all pairwise anchor-distance comparison events
\begin{align*}
\{h:\norm{h-a_i}_2^2\le \norm{h-a_j}_2^2\},
\qquad 1\le i,j\le N_t,
\end{align*}
with deterministic tie-breaking. The hard routing skeleton is measurable with respect to
\begin{align*}
\calF_t^{\mathrm{CA}}=
\calF_t^{\mathrm{edge}}\vee \calF_t^{\mathrm{anchor\text{-}order}}.
\end{align*}
The exact CA route map can be written as
\begin{align*}
I_{\calS_t}(h)
=
\left(
(g_e(h))_{e\in E_t},
\operatorname{HardRoute}_{\calS_t}(h)
\right).
\end{align*}
This map is measurable with respect to $\calF_t^{\mathrm{CA}}$, so
\begin{align*}
\sigma(I_{\calS_t})\subseteq \calF_t^{\mathrm{edge}}\vee \calF_t^{\mathrm{anchor\text{-}order}}.
\end{align*}
The inclusion can be strict: the full anchor-order sigma-algebra records pairwise order comparisons that may not affect the final selected concept. One may use $\calF_t^{\mathrm{CA}}$ as a stable upper information structure with simple repair-stability proofs, or use the coarser exact-route sigma-algebra $\sigma(I_{\calS_t})$ when the selected memory location itself is the desired state. Nodes are memory locations selected by this measurable route map; they need not be in one-to-one correspondence with atoms of the edge-skeleton partition. The soft read keeps the same anchored graph skeleton but may use movable centroids and smooth gates for locality; repairs can therefore change its effective smooth read geometry. Soft-CA should be read as a differentiable approximation to the anchor-generated hard information map, with exact partition claims recovered only in the hard-route limit or under explicit approximation assumptions.

\begin{theorem}[Anchor-based hard routing is graph-plus-anchor-order measurable, informal version of Theorem~\ref{thm:hard-dgm-measurable:formal}]
\label{thm:hard-dgm-measurable:informal}
With finite $E_t$, finite frozen anchors $a_1,\ldots,a_{N_t}$, and deterministic tie-breaking, the exact hard-route selected concept index, defined as the nearest eligible anchor with fallback to the nearest anchor, is measurable with respect to $\calF_t^{\mathrm{CA}}=\calF_t^{\mathrm{edge}}\vee \calF_t^{\mathrm{anchor\text{-}order}}$.
\end{theorem}

The next statement is analytic rather than structural: it applies to whichever fixed distances a soft read uses. With anchor distances it recovers the exact hard route; with centroid distances it describes the behavior of the soft retrieval approximation. It also exposes the candidate-recall requirement that links the formal hard route to the sparse implementation.

\begin{condition}[Candidate recall]
\label{cond:candidate-recall}
Let $j^\star(h)$ be the unique concept selected by the exact hard route for query $h$. The sparse soft read satisfies candidate recall at $h$ if
\begin{align*}
j^\star(h)\in N(h).
\end{align*}
If this condition fails, the soft read has no guarantee of selecting the hard route, regardless of how sharp the edge gates are.
\end{condition}

\begin{theorem}[Soft routing approaches hard routing, informal version of Theorem~\ref{thm:soft-routing-hard-limit:formal}]
\label{thm:soft-routing-hard-limit:informal}
Assume candidate recall holds for $h$, and let $j^\star$ be the recalled hard-route concept. Split the remaining candidates into eligible competitors, which satisfy the hard structural constraints, and wrong-side competitors, which violate at least one structural edge. Eligible competitors are removed by distance: if $j^\star$ has distance margin $\delta>0$ over every eligible competitor and the eligible/reference compatibility ratios are uniformly bounded, their total soft mass decays exponentially as $T\to0$. Wrong-side competitors are removed by edge gates: if every wrong-side candidate violates some incident structural edge with weight at least $\beta_{\min}>0$, signed margin at least $\delta_e>0$, and degree at most $D_{\max}$, then its compatibility decays uniformly after degree normalization. The hard-route limit is therefore an ordered or controlled joint limit: edge temperature must suppress wrong-side candidates before distance temperature makes nearest-neighbor selection decisive. Equivalently, either the wrong-side/reference distance-score ratios remain uniformly bounded, or for distance advantage $\Delta_d$ of wrong-side candidates over an eligible reference,
\begin{align*}
\frac{\beta_{\min}\delta_e}{\sqrt{D_{\max}}\tau_e}-\frac{\Delta_d}{T}\to+\infty.
\end{align*}
Under either condition, total wrong-side routing mass vanishes. Together with the distance margin for eligible competitors, this gives convergence to the hard route. Without candidate recall, the theorem gives no correctness guarantee for an omitted hard-route concept.
\end{theorem}

\begin{algorithm}[H]
\caption{\textsc{CA-Predict}}
\label{alg:dgm-predict}
\begin{algorithmic}[1]
\Statex $\calS_t=(\{a_j,c_j,M_j,n_j\}_{j=1}^{N_t},G_t),\quad x\in\calX$
\State $h\gets\phi_\theta(x)$
\State $N(h)\gets N_{k_c}^{\mathrm{centroid}}(h)\cup N_{k_a}^{\mathrm{anchor}}(h)\cup\operatorname{GraphNeighbors}(N_{k_c}^{\mathrm{centroid}}(h))$
\State $\chi_j(h)\gets
\exp\left\{
\frac{1}{\sqrt{\max\{1,\abs{\operatorname{Inc}_t(j)}\}}}
\sum_{e\in \operatorname{Inc}_t(j)}
\beta_e\log \operatorname{sigmoid}\left(s_{e,j}(u_e^\top h-b_e)/\tau_e\right)
\right\},\quad j\in N(h)$
\State $\alpha_j(h)\gets
\dfrac{\exp(-\norm{h-c_j}_2^2/T)\chi_j(h)}
{\sum_{r\in N(h)}\exp(-\norm{h-c_r}_2^2/T)\chi_r(h)},
\quad j\in N(h)$
\State $\hat a_t(h)\gets\sum_{j\in N(h)}\alpha_j(h)M_j(h)$
\State \Return $\hat a_t(h)$
\end{algorithmic}
\end{algorithm}

\subsection{Refine or repair on a graph}

After observing consequence $y$, CA scores empirical versions of the two gaps in Theorem~\ref{thm:error-decomposition}. A repair candidate keeps the CA information map fixed and updates only local state. A refinement candidate changes the information map by adding an anchored concept and distinction edges. The committed operation is the one whose held-out gain justifies its cost and safety margin.

Let $j^\star$ be the highest-weight concept under \eqref{eq:edge-gated-routing}. Define local surprise
\begin{align*}
\rho_t=L(y,M_{j^\star}(h)).
\end{align*}
Define neighborhood consequence disagreement on $N(h)$ by the task-appropriate local prediction divergence. For classification one may use
\begin{align*}
D_t=\max_{i,r\in N(h)}\operatorname{TV}(p_i(\cdot\mid h),p_r(\cdot\mid h)),
\end{align*}
where $p_j$ is the distribution represented by memory $M_j$. For regression one may use
\begin{align*}
D_t=\max_{i,r\in N(h)}\norm{\mu_i(h)-\mu_r(h)}_2,
\end{align*}
and for language modeling the log-loss surprise $-\log p_{j^\star}(x_{t+1}\mid x_{\le t})$ is the local disagreement signal. These quantities are proposal signals. They decide which candidates are worth scoring; they do not define repair or refinement.

All gap estimates below are computed on a held-out scoring buffer $\calB_t^{\mathrm{score}}$. Refinement proposals may use a disjoint proposal buffer $\calB_t^{\mathrm{prop}}$ together with the current observation; the scoring buffer is held out from proposal generation and from previous accept/reject decisions that determine the current state. Theorem~\ref{thm:safe-refinement:informal} applies when each scoring decision uses fresh held-out samples, or samples not previously used to choose proposals or update the state. Reservoir sampling can maintain disjoint proposal and scoring streams, but reusing the same validation buffer adaptively requires reusable-holdout or adaptive-validation control \citep{dwork2015reusable,dwork2015generalization}. Without that extra control, the empirical rule is a heuristic rather than the Hoeffding guarantee.

The edges created by a refinement are attached to a concrete confused-neighbor set. Let $\operatorname{Div}_{\calY}(M_i,M_r;h)$ denote a task-appropriate consequence divergence, such as total variation for classifiers, Euclidean mean distance for regression memories, or next-token distribution divergence for language models. Define
\begin{align*}
\operatorname{Confused}(h,y)
=
\left\{
i\in N(h):
L(y,M_i(h))>\tau_{\mathrm{conflict}}
\ \text{or}\
\operatorname{Div}_{\calY}(M_i,M_{j^\star};h)>\tau_{\mathrm{disagree}}
\right\}.
\end{align*}
When a budget of at most $q$ new edges is used, CA takes $\operatorname{Confused}_q(h,y)$ to be the $q$ highest-scoring members of this set under
\begin{align*}
\operatorname{score}_i(h,y)
=
\alpha_i(h)L(y,M_i(h))
+\gamma_{\mathrm{div}}\operatorname{Div}_{\calY}(M_i,M_{j^\star};h).
\end{align*}
If the set is empty but refinement is otherwise accepted, the implementation falls back to the highest-score candidates in $N(h)\setminus\{j^\star\}$. Thus ``confused neighbors'' denotes a concrete operator rather than an informal phrase.

Repair is guarded rather than blindly committed. The implementation proposes a small set of repair candidates, such as a memory-only update, a memory-plus-centroid update, and one or more reduced centroid steps:
\begin{align*}
\calC_{\mathrm{repair}}(\calS_t,h,y)
=
\{\calS_t^{\mathrm{mem}},\calS_t^{\mathrm{mem+centroid}},\calS_t^{\mathrm{shrink}},\calS_t\}.
\end{align*}
Each candidate keeps anchors and edges fixed, so it preserves $I_{\calS_t}$ and $\calF_t^{\mathrm{CA}}$. On the scoring buffer, the empirical repair gain is
\begin{align*}
\widehat\Delta_{\mathrm{repair}}
=
\widehat R(\calS_t;\calB_t^{\mathrm{score}})
-
\min_{\calS'\in\calC_{\mathrm{repair}}}
\widehat R(\calS';\calB_t^{\mathrm{score}}).
\end{align*}
CA's guarded repair candidate is
\begin{align}
\calS_{t+1}\in
\argmin_{\calS'\in\calC_{\mathrm{repair}}}
\widehat R(\calS';\calB_t^{\mathrm{score}})
\quad\text{subject to}\quad
\widehat R(\calS';\calB_t^{\mathrm{score}})
\le
\widehat R(\calS_t;\calB_t^{\mathrm{score}}).
\label{eq:guarded-repair}
\end{align}
The identity state is in $\calC_{\mathrm{repair}}$, so $\widehat\Delta_{\mathrm{repair}}\ge0$. Thus monotone repair is an enforced algorithmic property, not merely a theorem assumption. The anchor $a_{j^\star}$ is never moved by repair. Consequently, local memory repair and centroid adaptation do not rewrite previously accepted edge generators.

For a triggered refinement, define the candidate index $j_{\mathrm{new}}=N_t+1$ and the proposed edge set
\begin{align*}
E_{\mathrm{new}}(h,y)
=
\left\{
(j_{\mathrm{new}},i):
i\in\operatorname{Confused}_q(h,y),\
\norm{h-a_i}_2>\eta_{\mathrm{anchor}}
\right\}.
\end{align*}
The corresponding candidate state creates
\begin{align*}
a_{j_{\mathrm{new}}}=h,\qquad
c_{j_{\mathrm{new}}}=h,\qquad
M_{j_{\mathrm{new}}}=\operatorname{InitMemory}(h,y),
\end{align*}
with edge set $E_t\cup E_{\mathrm{new}}(h,y)$. Let $\calS_t^{\mathrm{new}}$ denote this state, and let $\widehat R(\calS;\calB_t^{\mathrm{score}})$ be the empirical local loss on the scoring buffer after routing with state $\calS$. The empirical penalized refinement gain is
\begin{align*}
\widehat\Delta_{\mathrm{refine}}
=\widehat R(\calS_t;\calB_t^{\mathrm{score}})
-
\widehat R(\calS_t^{\mathrm{new}};\calB_t^{\mathrm{score}})
-
\lambda_{\mathrm{concept}}
-
\lambda_{\mathrm{edge}}\abs{E_{\mathrm{new}}(h,y)}.
\end{align*}
Here $\lambda_{\mathrm{concept}}$ is the penalty for creating the anchored concept and its induced anchor-order bookkeeping, while $\lambda_{\mathrm{edge}}\abs{E_{\mathrm{new}}(h,y)}$ accounts for the accepted edge generators. Using the edge budget $q$ instead of the actual $\abs{E_{\mathrm{new}}(h,y)}$ is a conservative implementation choice that only makes the acceptance test stricter. The candidate is accepted only if it beats the repair gain by a safety margin:
\begin{align}
\widehat\Delta_{\mathrm{refine}}>
\max\{\widehat\Delta_{\mathrm{repair}},0\}
+\tau+4\epsilon_m
\qquad\text{and}\qquad
D_t>\tau_{\mathrm{disagree}}.
\label{eq:refine-accept}
\end{align}
Here $\epsilon_m$ is a statistical safety margin for a scoring buffer of size $m=\abs{\calB_t^{\mathrm{score}}}$; it can be set to zero for a purely empirical implementation, or to the adaptive validation margin in Theorem~\ref{thm:safe-refinement:informal}. If the refinement test fails, CA commits the guarded repair candidate in \eqref{eq:guarded-repair}. This makes refinement a statistically guarded gap comparison rather than an unscored growth heuristic.

\begin{figure}[t]
\centering
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[x=1cm,y=1cm,>=stealth]
\tikzset{
    stage/.style={draw,rounded corners=2pt,fill=gray!8,minimum width=1.75cm,minimum height=0.64cm,font=\scriptsize,align=center},
    repair/.style={draw,rounded corners=2pt,fill=green!10,minimum width=2.05cm,minimum height=0.72cm,font=\scriptsize,align=center},
    refine/.style={draw,rounded corners=2pt,fill=blue!8,minimum width=2.05cm,minimum height=0.72cm,font=\scriptsize,align=center},
    test/.style={draw,rounded corners=2pt,fill=orange!10,minimum width=2.65cm,minimum height=1.05cm,font=\scriptsize,align=center},
    commit/.style={draw,rounded corners=2pt,fill=purple!8,minimum width=1.75cm,minimum height=0.62cm,font=\scriptsize,align=center},
    lab/.style={font=\scriptsize}
}
\node[stage] (obs) at (0,0) {observe\\$(h_t,y_t)$};
\node[stage] (signals) at (2.05,0) {proposal signals\\$\rho_t,D_t,N(h_t)$};

\node[repair] (repaircand) at (4.55,0.9) {repair candidates\\fixed anchors and edges};
\node[refine] (refinecand) at (4.55,-0.9) {refinement candidate\\new anchor and edges};
\node[repair] (repairscore) at (7.15,0.9) {score on $\calB_t^{\mathrm{score}}$\\$\widehat\Delta_{\mathrm{repair}}$};
\node[refine] (refinescore) at (7.15,-0.9) {score on $\calB_t^{\mathrm{score}}$\\$\widehat\Delta_{\mathrm{refine}}$};
\node[test] (accept) at (10.0,0) {accept refinement if\\$\widehat\Delta_{\mathrm{refine}}>\widehat\Delta_{\mathrm{repair}}+\tau+4\epsilon_m$\\and $D_t>\tau_{\mathrm{disagree}}$};
\node[commit] (refinecommit) at (12.95,0.78) {commit\\refinement};
\node[commit] (repaircommit) at (12.95,-0.78) {commit\\guarded repair};

\draw[->,thick] (obs) -- (signals);
\draw[->,thick] (signals.east) -- (repaircand.west);
\draw[->,thick] (signals.east) -- (refinecand.west);
\draw[->,thick] (repaircand) -- (repairscore);
\draw[->,thick] (refinecand) -- (refinescore);
\draw[->,thick] (repairscore.east) -- (accept.north west);
\draw[->,thick] (refinescore.east) -- (accept.south west);
\draw[->,thick] (accept.north east) -- (refinecommit.west);
\draw[->,thick] (accept.south east) -- (repaircommit.west);
\end{tikzpicture}%
}
\caption{CA refine-or-repair decision. Proposal signals only choose candidates to score; the committed update is selected by held-out empirical gains. A refinement changes the information structure only when its penalized gain beats the best non-worsening repair by the safety margin; otherwise CA repairs local state while preserving anchors and edges.}
\label{fig:dgm-decision}
\end{figure}

\begin{algorithm}[H]
\caption{\textsc{CA-Observe}}
\label{alg:dgm-observe}
\begin{algorithmic}[1]
\Statex $\calS_t,\quad (x_t,y_t),\quad \calB_t^{\mathrm{prop}},\quad \calB_t^{\mathrm{score}}$
\State $h_t\gets\phi_\theta(x_t)$
\If{$N_t=0$}
\State $\calS_{t+1}\gets(\{(h_t,h_t,\operatorname{InitMemory}(h_t,y_t),1)\},(\{1\},\varnothing))$
\State \Return $\calS_{t+1}$
\EndIf
\State $\mathcal N_t\gets N_{k_c}^{\mathrm{centroid}}(h_t)\cup N_{k_a}^{\mathrm{anchor}}(h_t)\cup\operatorname{GraphNeighbors}(N_{k_c}^{\mathrm{centroid}}(h_t))$
\State $\alpha_{t,j}\gets\alpha_j(h_t),\quad j\in \mathcal N_t$
\State $j_t^\star\gets\argmax_{j\in \mathcal N_t}\alpha_{t,j}$
\State $\rho_t\gets L(y_t,M_{j_t^\star}(h_t))$
\State $D_t\gets\max_{i,r\in \mathcal N_t}\operatorname{Div}_{\calY}(M_i,M_r;h_t)$
\State $\calC_t^{\mathrm{repair}}\gets\calC_{\mathrm{repair}}(\calS_t,h_t,y_t)$
\State $\widehat\Delta_t^{\mathrm{repair}}\gets
\widehat R(\calS_t;\calB_t^{\mathrm{score}})
-\min_{\calS'\in\calC_t^{\mathrm{repair}}}
\widehat R(\calS';\calB_t^{\mathrm{score}})$
\State $\calS_t^{\mathrm{repair}}\in
\argmin_{\substack{\calS'\in\calC_t^{\mathrm{repair}}\\
\widehat R(\calS';\calB_t^{\mathrm{score}})
\le \widehat R(\calS_t;\calB_t^{\mathrm{score}})}}
\widehat R(\calS';\calB_t^{\mathrm{score}})$
\State $z_t\gets\one\{\rho_t>\tau_{\mathrm{repair}}\ \vee\ D_t>\tau_{\mathrm{disagree}}\}$
\If{$z_t=1$}
\State $E_t^{\mathrm{new}}\gets E_{\mathrm{new}}(h_t,y_t),\quad \calS_t^{\mathrm{new}}\gets\operatorname{AddConcept}(\calS_t,h_t,y_t,E_t^{\mathrm{new}};\calB_t^{\mathrm{prop}})$
\State $\widehat\Delta_t^{\mathrm{refine}}\gets
\widehat R(\calS_t;\calB_t^{\mathrm{score}})
-\widehat R(\calS_t^{\mathrm{new}};\calB_t^{\mathrm{score}})
-\lambda_{\mathrm{concept}}
-\lambda_{\mathrm{edge}}\abs{E_t^{\mathrm{new}}}$
\State $r_t\gets
\one\{\widehat\Delta_t^{\mathrm{refine}}>
\max\{\widehat\Delta_t^{\mathrm{repair}},0\}+\tau+4\epsilon_m
\ \wedge\ D_t>\tau_{\mathrm{disagree}}\}$
\If{$r_t=1$}
\State $\calS_{t+1}\gets\calS_t^{\mathrm{new}}$
\State \Return $\calS_{t+1}$
\EndIf
\EndIf
\State $\calS_{t+1}\gets\calS_t^{\mathrm{repair}}$
\State \Return $\calS_{t+1}$
\end{algorithmic}
\end{algorithm}

\begin{definition}[Guarded monotone local repair]
\label{def:monotone-repair}
A guarded repair operator is monotone on scoring buffer $\calB_t^{\mathrm{score}}$ if it commits a candidate state $\calS_{t+1}$ only when local memories and retrieval centroids may change, anchors and edges remain fixed, and
\begin{align*}
\widehat R(\calS_{t+1};\calB_t^{\mathrm{score}})
\le
\widehat R(\calS_t;\calB_t^{\mathrm{score}}).
\end{align*}
The identity candidate $\calS_{t+1}=\calS_t$ is always allowed, so the guard is feasible even when every nontrivial repair would hurt the scoring buffer.
\end{definition}

\begin{theorem}[Refine-or-repair objective descent, informal version of Theorem~\ref{thm:refine-repair-descent:formal}]
\label{thm:refine-repair-descent:informal}
Let the local scoring objective on a fixed scoring buffer $\calB_t^{\mathrm{score}}$ be
\begin{align*}
J_{\calB_t^{\mathrm{score}}}(\calS)
=
\widehat R(\calS;\calB_t^{\mathrm{score}})
+\lambda_{\mathrm{concept}}N_{\mathrm{concept}}(\calS)
+\lambda_{\mathrm{edge}}\abs{E(\calS)}.
\end{align*}
If repair is committed by the guarded rule \eqref{eq:guarded-repair}, then $J_{\calB_t^{\mathrm{score}}}$ does not increase. If a refinement adding one anchored concept and actual edge set $E_{\mathrm{new}}$ is accepted only under \eqref{eq:refine-accept}, then the empirical penalized objective $J_{\calB_t^{\mathrm{score}}}$ decreases by more than $\widehat\Delta_{\mathrm{repair}}+\tau+4\epsilon_m$ and in particular by more than $\tau+4\epsilon_m$.
\end{theorem}

This is a local-buffer statement. It does not imply monotone descent of a global objective when the scoring buffer changes over time; a global finiteness claim requires a common potential or validation objective.

\begin{theorem}[Statistically safe adaptive refinement, informal version of Theorem~\ref{thm:safe-refinement:formal}]
\label{thm:safe-refinement:informal}
Assume losses lie in $[0,1]$. Let $(\mathfrak H_t)$ be the filtration containing the past stream, the current CA state, and all proposal-buffer information available before scoring at time $t$. Let $\calC_t$ be a finite candidate class that is predictable with respect to $\mathfrak H_t$, and let $\calB_t^{\mathrm{score}}$ contain $m_t$ fresh held-out samples, or samples not previously used to choose proposals or update the state, independent of $\mathfrak H_t$ conditional on the current scoring distribution. For candidate $g$, let $\widehat G_t(g)$ be the empirical raw risk decrease on $\calB_t^{\mathrm{score}}$ before complexity cost, and let $G_t(g)$ be the corresponding population raw gain. Define
\begin{align*}
\epsilon_t=
\sqrt{
\frac{\log(4t^2(\abs{\calC_t}+1)/\delta)}
{2m_t}
}.
\end{align*}
Then, with probability at least $1-\delta$, for every time $t$ and every $g\in\calC_t$,
\begin{align*}
\abs{\widehat G_t(g)-G_t(g)}
\le
2\epsilon_t.
\end{align*}
Hence an accepted refinement satisfying
\begin{align*}
\widehat G_t(g)>
\lambda\operatorname{cost}(g)+\tau+2\epsilon_t
\end{align*}
equivalently $\widehat\Delta_t^{\mathrm{pen}}(g)=\widehat G_t(g)-\lambda\operatorname{cost}(g)>\tau+2\epsilon_t$, has positive population penalized gain after paying complexity cost and margin $\tau$, uniformly over all accepted times.
\end{theorem}

\begin{corollary}[Bounded language-model refinement score, informal version of Corollary~\ref{cor:lm-clipped-refinement:formal}]
\label{cor:lm-clipped-refinement:informal}
For language modeling, raw next-token negative log likelihood is unbounded. Fix $L_{\max}>0$ and define the clipped normalized score
\begin{align*}
\ell_{\mathrm{clip}}(x_{\le t},x_{t+1})
=
\frac{1}{L_{\max}}
\min\{-\log p(x_{t+1}\mid x_{\le t}),L_{\max}\}.
\end{align*}
If refinement proposal and scoring use $\ell_{\mathrm{clip}}$, then Theorem~\ref{thm:safe-refinement:informal} applies to clipped-loss refinement gain. An accepted refinement has positive clipped population gain after cost and margin with the stated high probability. This corollary does not claim positive gain for the raw unbounded negative log likelihood without additional tail assumptions.
\end{corollary}

\begin{theorem}[Anchor-stable CA information structure, informal version of Theorem~\ref{thm:anchor-stable:formal}]
\label{thm:anchor-stable:informal}
Let
\begin{align*}
\calF_t^{\mathrm{CA}}
=
\sigma(g_e:e\in E_t)
\vee
\sigma\left(\{h:\norm{h-a_i}_2^2\le \norm{h-a_j}_2^2\}:1\le i,j\le N_t\right).
\end{align*}
If a repair step changes only retrieval centroids $c_j$, local memories $M_j$, and counts $n_j$, while keeping anchors $a_j$ and edges $E_t$ fixed, then $\calF_{t+1}^{\mathrm{CA}}=\calF_t^{\mathrm{CA}}$. If a refinement adds a new anchor and new edge generators, then the updated information structure is the join of $\calF_t^{\mathrm{CA}}$ with the new anchor-order events and the new edge generators.
\end{theorem}

Thus the edge skeleton realizes the minimal join operation from Section~\ref{sec:minimal-refinement}. The exact hard-route CA adds anchor-order events as routing infrastructure whenever a new anchor is inserted. Consequently, Hard-CA is a structured implementation of refinement, but its exact-route information update is not always the single-event minimal refinement in the strict Section~\ref{sec:minimal-refinement} sense.

\begin{corollary}[Finite number of accepted refinements, informal version of Corollary~\ref{cor:finite-refinements:formal}]
\label{cor:finite-refinements:informal}
If there is a common nonnegative potential $\Phi(\calS)$ and every accepted refinement decreases $\Phi$ by at least $\tau>0$, then the number of accepted refinements is at most $\Phi(\calS_0)/\tau$. In particular, this applies to a fixed scoring buffer or validation objective. The local-buffer descent theorem alone should not be read as a global finiteness result when the scoring buffer changes over time.
\end{corollary}

\subsection{Data-structure operations}

\begin{definition}[CA data-structure operations]
\label{def:dgm-data-structure-operations}
Viewed as a dynamic partition data structure, CA implements four operations:
\begin{align*}
\operatorname{Query}(h),\qquad
\operatorname{Repair}(h,y),\qquad
\operatorname{Refine}(h,y,g),\qquad
\operatorname{Score}(g,\calB).
\end{align*}
The maintained edge-skeleton invariant is
\begin{align*}
\operatorname{Repair}\quad&\Longrightarrow\quad \calP^G(\calS_{t+1})=\calP^G(\calS_t),\\
\operatorname{Refine}(g)\quad&\Longrightarrow\quad \calP^G(\calS_{t+1})=\calP^G(\calS_t)\vee\sigma(g),
\end{align*}
where $\calP^G(\calS_t)$ is the edge-generated skeleton partition. The exact hard-route invariant is
\begin{align*}
\operatorname{Repair}\quad&\Longrightarrow\quad \calF^{\mathrm{CA}}(\calS_{t+1})=\calF^{\mathrm{CA}}(\calS_t),\\
\operatorname{Refine}\quad&\Longrightarrow\quad \calF^{\mathrm{CA}}(\calS_{t+1})
=
\calF^{\mathrm{CA}}(\calS_t)\vee\sigma(g_e:e\in E_{\mathrm{new}})\vee\sigma(\text{new anchor-order events}).
\end{align*}
Thus repair modifies payloads on the current structure, while refinement changes the structure by adding both edge generators and the order information induced by new anchors.
\end{definition}

\begin{theorem}[CA data-structure cost, informal version of Theorem~\ref{thm:dgm-complexity:formal}]
\label{thm:dgm-data-structure-main}
Assume $N_t$ concepts in $\R^d$, candidate set size $\abs{N(h)}=k$, selected incident edge count $Q(h)$, retrieval cost $C_{\mathrm{ret}}(N_t,d)$, and selected local-memory evaluation cost $C_{\mathrm{mem}}$. Then
\begin{align*}
\operatorname{Query}(h)
=
O(C_{\mathrm{ret}}(N_t,d)+Q(h)d+kd+C_{\mathrm{mem}}).
\end{align*}
Here $Q(h)$ is an edge count and each dense linear edge gate costs $O(d)$ to evaluate. A repair adds the cost of scoring the proposed local-memory candidates on the scoring buffer. A refinement creating one anchor-centroid pair and $q$ guarded linear distinction edges adds $O(d+qd)$ state-update time, plus the indexing or comparison bookkeeping used to maintain anchor-order queries. With sparse categorical memories of average support $\bar s$, the state uses
\begin{align*}
O(N_t(2d+\bar s)+\abs{E_t}d)
\end{align*}
memory, up to indexing overhead.
\end{theorem}

The theorem is an implementation bound rather than a statistical guarantee. It says that CA realizes the abstract operation $\Pi\mapsto\Pi\vee U$ on the edge skeleton and the corresponding join update on $\calF^{\mathrm{CA}}$ with local graph updates and sparse routing, provided the candidate retrieval layer supplies the relevant hard-route concepts.

\paragraph{Comparison to gradient and memory baselines.}
Backpropagation is not itself a repair or refinement operation; it is an optimizer. If it updates only parameters downstream of a fixed information map, it is a fixed-information repair update in the sense of Definition~\ref{def:fixed-information-repair}. If it updates the encoder or routing representation, it may change the induced partition and is outside the repair-only lower bound. Table~\ref{tab:runtime-memory-comparison} summarizes the resulting computational comparison against common online optimizers, caches, trees, and parameter-efficient adaptation modules \citep{rumelhart1986learning,kingma2015adam,grave2017improving,khandelwal2020generalization,domingos2000mining,houlsby2019parameter,hu2022lora}. Let $P_{\mathrm{tr}}$ be the number of trainable parameters updated by backpropagation, $A_{\mathrm{bp}}$ the activation storage required for the backward pass over the adaptation window, $M$ the number of stored examples in a cache or $k$NN memory, $D_{\mathrm{tree}}$ the tree depth, $S_{\mathrm{leaf}}$ the per-leaf sufficient statistics, and $\bar s$ the average sparse categorical support per CA concept. The table lists runtime-adaptation overhead beyond any shared frozen forward pass that computes $h$.

\begin{table}[t]
\centering
\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.18\linewidth}
>{\raggedright\arraybackslash}p{0.21\linewidth}
>{\raggedright\arraybackslash}p{0.22\linewidth}
>{\raggedright\arraybackslash}p{0.23\linewidth}}
\toprule
Method & Information operation & Per-round adaptation / read overhead & Runtime state memory \\
\midrule
Full backprop + Adam & May change information map if encoder/routing parameters move & Backward/update through $P_{\mathrm{tr}}$ parameters; activation storage for the adaptation window & Weights, gradients, and Adam moments: $O(P_{\mathrm{tr}})$ each, plus $O(A_{\mathrm{bp}})$ activations \\
Head-only or linear-probe backprop & Fixed-information repair when features $h=I(\omega)$ are frozen & Backward/update through head parameters $P_{\mathrm{head}}$ & $O(P_{\mathrm{head}})$ weights plus optimizer state and head activations \\
Adapters / LoRA online & Repair if downstream of fixed $I$; otherwise may change the information map & Backward/update through adapter parameters $P_{\mathrm{ad}}$ in selected layers & $O(P_{\mathrm{ad}})$ weights plus optimizer state and selected-layer activations \\
Cache or budgeted $k$NN & Stores examples under a fixed representation; no explicit refinement test & Retrieval $C_{\mathrm{ret}}(M,d)$ plus aggregation over $k$ neighbors & $O(M(d+\bar s))$ keys and values, plus index overhead \\
Online tree / Hoeffding tree & Refines a rooted leaf partition & Query $O(D_{\mathrm{tree}})$; update split statistics on visited leaf & $O(\#\mathrm{leaves}\cdot S_{\mathrm{leaf}})$ \\
CA, exact Hard-CA & Refines graph-generated partition by adding binary generators & Query $O(N_td+Q(h)d+kd+C_{\mathrm{mem}})$ with exact retrieval; accepted refinement adds $O(d+qd)$ & $O(N_t(2d+\bar s)+\abs{E_t}d)$ \\
CA, sparse Soft-CA & Approximate Hard-CA under candidate recall and margins & Query $O(C_{\mathrm{ret}}(N_t,d)+Q(h)d+kd+C_{\mathrm{mem}})$; accepted refinement adds $O(d+qd)$ & $O(N_t(2d+\bar s)+\abs{E_t}d)$ plus index overhead \\
\bottomrule
\end{tabular}
\caption{Runtime-adaptation cost comparison. CA does not claim to dominate unconstrained full backpropagation statistically. The theoretical separation applies to fixed-information repair methods: when the representation quotient is fixed and has positive refinement gap, they incur the lower bound in Theorem~\ref{thm:repair-only-regret}, while CA can pay local graph-memory cost to add a distinction. Full backpropagation through the representation can also change the quotient, but pays backward-pass, activation, and optimizer-state costs.}
\label{tab:runtime-memory-comparison}
\end{table}

\subsection{Canonical local memory}

\begin{definition}[Calibrated categorical local memory]
\label{def:calibrated-categorical-memory}
The local memory $M_j$ is the repair mechanism inside an already chosen concept. For classification and language modeling, the main-text canonical memory is a calibrated categorical memory rather than an unconstrained embedding score, consistent with proper probabilistic scores and calibration diagnostics for probability forecasts \citep{brier1950verification,gneiting2007strictly,guo2017calibration}. Concept $j$ stores token or label counts $n_{j,y}$ and $n_j=\sum_y n_{j,y}$. Given a backbone distribution $p_0(y\mid h)$ and smoothing strength $\alpha_0>0$, define
\begin{align*}
p_j(y\mid h)
=
\frac{n_{j,y}+\alpha_0 p_0(y\mid h)}
{n_j+\alpha_0}.
\end{align*}
The routed memory distribution is
\begin{align*}
p_{\mathrm{mem}}(y\mid h)
=
\sum_{j\in N(h)}\alpha_j(h)p_j(y\mid h),
\end{align*}
and the final prediction is the convex mixture
\begin{align*}
p_{\mathrm{CA}}(y\mid h)
=
(1-\lambda_t)p_0(y\mid h)+\lambda_t p_{\mathrm{mem}}(y\mid h),
\qquad
\lambda_t\in[0,1].
\end{align*}
\end{definition}
This keeps the read path a valid probability distribution. It also gives the simple pointwise safety bound
\begin{align*}
-\log p_{\mathrm{CA}}(y\mid h)
\le
-\log p_0(y\mid h)-\log(1-\lambda_t)
\end{align*}
whenever $\lambda_t<1$. For log-loss experiments we use a stronger cumulative safeguard based on standard prediction-with-expert-advice and aggregating-algorithm bounds \citep{vovk1990aggregating,freund1997decision,cesabianchi2006prediction}. Let $\Lambda=\{\lambda_1,\ldots,\lambda_K\}\subseteq[0,1]$ with $0\in\Lambda$, define
\begin{align*}
q_{t,k}(y)
=
(1-\lambda_k)p_0(y\mid h_t)+\lambda_k p_{\mathrm{mem}}(y\mid h_t),
\end{align*}
and predict with a mixture $p_t(y)=\sum_k w_{t,k}q_{t,k}(y)$. After observing $y_t$, update the trust weights by
\begin{align}
w_{t+1,k}
=
\frac{w_{t,k}q_{t,k}(y_t)}
{\sum_{\ell=1}^K w_{t,\ell}q_{t,\ell}(y_t)}.
\label{eq:trust-weight-update}
\end{align}

\begin{theorem}[No-harm log-loss trust aggregation, standard]
\label{thm:dgm-no-harm-log-loss}
Initialize $w_{1,k}=1/K$ and update by \eqref{eq:trust-weight-update}. For any outcome sequence,
\begin{align*}
\sum_{t=1}^T -\log p_t(y_t)
\le
\min_{k\in[K]}\sum_{t=1}^T -\log q_{t,k}(y_t)+\log K.
\end{align*}
Since $0\in\Lambda$, one expert is the frozen backbone predictor, so
\begin{align*}
\sum_{t=1}^T -\log p_t(y_t)
\le
\sum_{t=1}^T -\log p_0(y_t\mid h_t)+\log K.
\end{align*}
\end{theorem}

\begin{proof}
Let $Z_t=\sum_k w_{1,k}\prod_{s=1}^t q_{s,k}(y_s)$. The posterior update implies $p_t(y_t)=Z_t/Z_{t-1}$, hence $\prod_{t=1}^T p_t(y_t)=Z_T$. For every $k$, $Z_T\ge K^{-1}\prod_{t=1}^T q_{t,k}(y_t)$. Taking negative logarithms and minimizing over $k$ gives the first bound. The second follows from the $\lambda=0$ expert.
\end{proof}

Because $p_0(y\mid h)$, $M_j(h)$, and the soft weights can depend on the hidden state, the practical predictor need not be constant on atoms of the edge partition alone. The formal interpretation is the base-information convention from Section~\ref{sec:preliminaries}: let
\begin{align*}
\calB=\sigma\left(p_0(\cdot\mid h),\text{fixed pre-outcome features used outside CA routing}\right).
\end{align*}
Then CA refinements are claims about the additional runtime information in $\calF_t^{\mathrm{CA}}$, and predictors are measured as $\calB\vee\calF_t^{\mathrm{CA}}$-measurable. The hidden state may be used by the algorithm to compute admissible distinctions, proposal scores, and restricted routing functions, but it is not treated as arbitrary freely conditionable information in the unrestricted Bayes-risk theory. If full $h$ is admitted into $\calB$, then all edge predicates and anchor-order events are already $\calB$-measurable and the partition-level refinement gap collapses to zero. In the hard one-hot special case, where each memory returns an atomwise constant action and no extra base features are used, this reduces to strict partition measurability; $h$-dependent soft memories should be read as restricted-class or approximation statements.

Repair is the sufficient-statistic update. For every selected concept $j\in N(h)$,
\begin{align*}
n_{j,y}^+=n_{j,y}+\alpha_j(h),
\qquad
n_j^+=n_j+\alpha_j(h),
\end{align*}
with all unselected counts unchanged. The low-cost hard-update variant is recovered by setting $\alpha_{j^\star}=1$ for the highest-weight concept and $\alpha_j=0$ for all $j\ne j^\star$. Embedding-mean memories and ridge or projection-style local repair are useful implementation variants, but they should be evaluated with calibration metrics before being used for log-loss claims.

\subsection{From CA to CALM}
\label{sec:language-dgm}

The generic CA object above can be attached to many local memories. For sequence models, Appendix~\ref{sec:large-model-training} gives CALM as an attention-replacement architecture with a causal read-before-write protocol and detached runtime graph updates.
