\section{Evaluation Design: Inference-Time Adaptation}
\label{sec:experiments}

The empirical test of DGM should be as sharp as the theory: it must expose the
difference between repairing a fixed information structure and refining that
structure when the current state has collapsed consequence-relevant cases. We
therefore use a prequential inference-time protocol \citep{dawid1984prequential}
with a shared frozen backbone. All mandatory runs are sized for one Apple M4 Max
workstation: no distributed training, no server-side retrieval service, and no
offline hyperparameter search that would be unavailable to the compared methods
at runtime.

\subsection{Runtime contract}
\label{sec:runtime-contract}

Every method receives the same frozen encoder or language model. The cost of
training or loading this shared model is reported but not charged to runtime
adaptation. At round $t$, the method receives $x_t$, computes the shared hidden
state $h_t=\phi_\theta(x_t)$, predicts $\hat y_t$, then observes the consequence
$y_t$ and may update only its declared runtime state:
\[
x_t \longrightarrow h_t \longrightarrow \hat y_t,
\qquad
y_t \longrightarrow \text{runtime update}.
\]
DGM updates by detached graph-memory repair or by guarded refinement. Test-time
training baselines update a head, prompt, adapter, normalization state, or other
declared adaptation parameters by inference-time gradient steps. The reported
adaptation overhead excludes the shared frozen forward pass and includes all
extra reads, writes, backward passes, activation storage, optimizer state,
synchronization, and memory-index maintenance.

\subsection{Questions}
\label{sec:experiment-questions}

The suite is organized around four falsifiable questions.
\begin{enumerate}
    \item \textbf{Structural necessity.} On streams with a positive refinement
    gap, does full DGM reduce prequential loss where repair-only memories,
    caches, online ridge, and head-only backpropagation are stuck under the same
    frozen information map?
    \item \textbf{Non-gratuitous refinement.} On streams whose errors are
    repairable inside the current concepts, does DGM mostly repair rather than
    create unnecessary nodes and edges?
    \item \textbf{Natural frozen-model adaptation.} On image and language
    streams with revealed labels or next tokens, does DGM improve the
    loss--latency--memory frontier relative to inference-time gradient
    adaptation and nonparametric memory baselines?
    \item \textbf{Scaling on one workstation.} How far can concept count, edge
    count, stream length, and hidden dimension be pushed on an M4 Max before
    latency or unified-memory use becomes the limiting factor?
\end{enumerate}

\subsection{Benchmarks}
\label{sec:experiment-benchmarks}

\begin{table}[t]
\centering
\footnotesize
\begin{tabular}{>{\raggedright\arraybackslash}p{0.18\linewidth}
>{\raggedright\arraybackslash}p{0.28\linewidth}
>{\raggedright\arraybackslash}p{0.24\linewidth}
>{\raggedright\arraybackslash}p{0.18\linewidth}}
\toprule
Benchmark & What it tests & M4 Max scale target & Primary success condition \\
\midrule
Hidden-context control & The frozen state intentionally collapses regimes that require different predictions. The oracle refined partition is known. & $10^6$--$10^7$ stream rounds, $K\in\{8,32,128\}$ regimes, label noise grid, 5 seeds. & DGM approaches the refined oracle and beats repair-only regret slope. \\
Repair-only control & Same stream geometry but no hidden consequence conflict. Errors come from stale local statistics. & Matched stream length and concept budget. & DGM accepts few refinements and matches the best repair baseline. \\
Frozen vision streams & CIFAR-10/100-style frozen-encoder classification \citep{krizhevsky2009learning} with class/domain bursts, corruptions, and replay-free label reveal. & Full CIFAR-10 and CIFAR-100 streams; optional resized ImageNet-100 or local image subset if available. & Better prequential NLL or accuracy at matched adaptation time and memory. \\
Frozen LM streams & A frozen next-token model processes document continuations; the observed next token is the consequence. & $5$M--$50$M tokens with cached fp16 hidden states; char/word small model for smoke, GPT-style frozen model for max run. & Lower prequential NLL than cache and gradient-update adapters at lower backward cost. \\
Systems scaling & Synthetic hidden states plus labels isolate data-structure cost from model forward cost. & $N=10^3$--$10^6$ concepts, $d=128$--$2048$, top-$k=8$--$64$, edge degree $q=0$--$32$. & Report latency, memory, candidate recall, and refinement throughput. \\
\bottomrule
\end{tabular}
\caption{Evaluation suite. The controlled streams test the theorem-level
separation; the image and language streams test practical frozen-backbone
adaptation; the systems benchmark establishes the largest single-workstation
operating envelope.}
\label{tab:dgm-evaluation-suite}
\end{table}

\paragraph{Controlled hidden-context streams.}
The cleanest experiment makes the repair obstruction visible. A stream contains
a hidden regime $z_t$ and an observed feature $x_t$ whose frozen representation
collapses multiple regimes into the same current atom. Labels are generated so
that $P(y\mid x,z)$ differs across hidden regimes while $P(y\mid x)$ is mixed.
The refined oracle observes the regime bit or an equivalent distinction; repair
baselines do not. DGM receives only the frozen hidden state and consequences, so
it must discover distinction edges from repeated local contradictions. We report
regret slope against the refined oracle, accepted concepts and edges, and the
empirical mutual-information gain of accepted distinctions. A matched
repair-only control removes the hidden conflict while preserving marginal
difficulty; DGM should then repair without growing a large graph.

\paragraph{Frozen vision adaptation.}
The image benchmark uses a frozen encoder trained once on a disjoint pretraining
split or loaded from a fixed checkpoint. The adaptation stream is predict-then-
observe: labels arrive after each prediction. We use three stream orders:
class-balanced random order, class-burst order, and domain/corruption-burst
order. Class bursts test rapid local memory updates; domain bursts test whether
refinement creates consequence-relevant concepts instead of only appending
examples to a cache. The largest mandatory run uses all CIFAR-100 train examples
as the adaptation stream and the official test split as held-out evaluation;
CIFAR-10 is retained for fast ablations and debugging.

\paragraph{Frozen language-model adaptation.}
For language modeling, the consequence is the next observed token. The model
predicts token $x_{t+1}$ from prefix $x_{\le t}$, then uses the revealed token to
repair or refine DGM-Embedding. The main stream uses long contiguous documents
so that local entities, notation, formatting, and topic shifts create repeated
contexts. Free generations are not used as observations. To make the M4 Max run
large enough without conflating model forward cost with memory cost, the
largest run may cache frozen hidden states in fp16 and replay the prequential
adaptation layer over $5$M--$50$M tokens. A smaller end-to-end run recomputes
the frozen model forward pass to verify that the cached-hidden protocol matches
online execution.

\subsection{Baselines and ablations}
\label{sec:experiment-baselines}

The core baselines are deliberately close competitors rather than weak offline
training comparisons:
\begin{itemize}
    \item \textbf{Frozen predictor:} no runtime adaptation.
    \item \textbf{Repair-only memories:} calibrated class/token counts, online
    ridge/RLS, passive-aggressive updates, and CPM-style projection memory under
    the same frozen features.
    \item \textbf{Cache and $k$NN memories:} budgeted example stores with the
    same key dimension, retrieval budget, and memory cap as DGM.
    \item \textbf{Online tree baselines:} Hoeffding-style or Mondrian-style
    online partitions on the frozen hidden states, where applicable.
    \item \textbf{Inference-time gradient baselines:} one-step and multi-step
    updates to a linear head, soft prompt, adapter, LoRA block, or normalization
    state, with optimizer state and activation memory counted.
    \item \textbf{Full DGM:} guarded repair, candidate-recall retrieval,
    consequence-sensitive refinement, and edge-gated routing.
\end{itemize}
The mandatory DGM ablations are repair-only DGM, refine-only DGM, edge-free
DGM, no anchor-recall candidates, no graph-neighbor candidates, unguarded
refinement, and fixed trust weight versus online-selected trust weight. These
ablations identify whether gains come from refinement, edge gating, better
candidate recall, or merely extra memory capacity.

\subsection{Metrics and reporting}
\label{sec:experiment-metrics}

Each run reports prequential loss, prequential accuracy when applicable,
held-out post-adaptation loss, and regret to the known oracle on controlled
streams. Systems measurements are separated into shared forward time, adaptation
read time, adaptation write time, backward time, synchronization time, peak
unified-memory use, optimizer-state memory, activation memory, concept memory,
and index memory. We also report number of backward passes, trainable
adaptation parameters, accepted concepts, accepted edges, average degree,
refinement acceptance rate, candidate-recall rate for controlled streams,
calibration error for probabilistic outputs, and the validation gain of accepted
versus rejected refinements.

\subsection{Decision rule for claiming evidence}
\label{sec:experiment-claim-rule}

The experiment supports the DGM claim only if three outcomes hold together.
First, on hidden-context streams, full DGM must reduce the repair-only regret
slope and move toward the refined oracle with a graph whose accepted distinctions
have positive held-out value. Second, on repair-only controls, DGM must avoid
unnecessary growth and remain competitive with the best fixed-information
repair method. Third, on natural frozen-model streams, DGM must improve the
prequential loss--latency--memory frontier against cache and inference-time
gradient baselines. If DGM only wins by using more memory, by seeing labels
before prediction, or by hiding backbone/gradient cost, the experiment does not
validate the theory.
