\section{Guarantees}
\label{sec:guarantees}

The guarantees are deliberately scoped. They do not assert global optimality of CA, nor do they assert that a deep backbone can always discover the right representation. They formalize the paper's decision rule: repair-only learners are limited by the Bayes risk of their current partition; useful refinements change the Bayes action; statistically guarded refinement accepts only distinctions whose held-out consequence value clears complexity and concentration margins; and CA implements this operation as a dynamic graph data structure. Full proof details are in Appendices~\ref{app:refinement-proofs}, \ref{app:value-proofs}, and \ref{app:algorithm-bridge-proofs}.

\paragraph{1. Error decomposes into two gaps.}
Theorem~\ref{thm:error-decomposition} states that for current predictor $f_t$, current partition $\Pi_t$, and candidate refined partition $\Pi^\star$,
\begin{align*}
R_L(f_t)-R_L(\Pi^\star)
=
\Delta_{\mathrm{repair}}(f_t,\Pi_t)
+
\Delta_{\mathrm{refine}}(\Pi_t,\Pi^\star).
\end{align*}
The first term is the best possible gain from repairing the predictor at fixed information. The second is the value of changing the information structure.

\paragraph{2. Repair-only learners have linear regret.}
Theorem~\ref{thm:repair-only-regret} states that any learner whose predictions remain measurable with respect to a fixed partition $\Pi$ has expected regret at least
\begin{align*}
T\left(R_L(\Pi)-R_L(\Pi^\star)\right)
\end{align*}
against a refined oracle $\Pi^\star$. Thus a positive refinement gap is not just a pointwise contradiction; it is a sequential lower bound. Corollary~\ref{cor:fixed-information-gradient-repair} clarifies the relationship to backpropagation: gradient updates are repair-only only when they update parameters downstream of a fixed information map. Full backpropagation through the representation can change the information structure and is not covered by this lower bound.

\paragraph{3. Useful refinements change the Bayes action.}
Theorem~\ref{thm:loss-relevant-refinement:informal} states that, under uniqueness of the refined Bayes action, a refinement has zero value exactly when the refined action is already measurable under the old information, in the classical decision-theoretic sense of information value \citep{blackwell1953equivalent,berger1985statistical}. For finite $Y$ under log loss, Theorem~\ref{thm:log-value:informal} sharpens this to
\begin{align*}
R_{\log}(\calF)-R_{\log}(\calG)
=
\E D_{\mathrm{KL}}\left(
\mathbb P(Y\in\cdot\mid\calG)
\middle\|
\mathbb P(Y\in\cdot\mid\calF)
\right).
\end{align*}
Thus stochastic conflict, not only hard-label contradiction, is enough to make refinement valuable.

These value identities are population statements. Hoeffding-style validation results later in the paper require bounded loss, clipped loss, or separate tail assumptions; raw language-model negative log likelihood is handled only through the clipped-score corollary unless additional concentration assumptions are supplied.

\paragraph{4. Missing distinctions obstruct repair.}
Corollary~\ref{thm:contradiction:informal} says that if two states have the same information-map value but require different consequences, no predictor measurable with respect to the current information can satisfy both. This is the main theoretical point of the paper:
\begin{align*}
\boxed{\text{if the target separates states that the information map identifies, repair is impossible.}}
\end{align*}
The result is architecture-independent; it applies to caches, adapters, local memories, and test-time optimizers whenever their update leaves the relevant information structure fixed.

\paragraph{5. Refinement has a canonical minimal form.}
Theorems~\ref{thm:minimal-refinement:informal} and \ref{thm:sigma-refinement:informal} identify the least information update that makes new evidence observable. In the finite case, a binary distinction splits exactly the atoms it crosses. In the sigma-algebra case, the update is
\begin{align*}
\calF_{t+1}=\calF_t\vee\sigma(U_t).
\end{align*}
Section~\ref{sec:value} gives the standard value identities that explain when such a distinction is useful: conditional mutual information under log loss and conditional variance reduction under squared loss \citep{cover2006elements,gneiting2007strictly,billingsley1995probability,kallenberg2002foundations}. Theorem~\ref{thm:minimal-loss-sufficient:informal} further identifies the smallest refinement needed to express the refined Bayes action.

For CA specifically, the edge skeleton realizes this minimal join operation. The exact hard route also adds anchor-order events as routing infrastructure when a new anchor is inserted, so the full Hard-CA update is a structured refinement rather than always a single-event minimal refinement.

\paragraph{6. Greedy refinement has finite-class guarantees.}
Theorem~\ref{thm:greedy-recovery:informal} gives a realizable recovery guarantee: if the algorithm keeps selecting positive-mass impure atoms, each selected atom has a target-respecting distinction with value at least $\gamma$ that creates two positive-mass sides, and split values are estimated accurately enough, greedy refinement recovers the target partition on the positive-mass support after at most $K^+-m_0^+\le K^+-1$ accepted splits, where $K^+$ and $m_0^+$ count positive-mass target and initial atoms. Theorem~\ref{thm:oracle-refinement:informal} gives a finite-class oracle inequality over partitions reachable by at most $B$ global distinctions; the atom count in that class is $m_0 2^B$, not $m_0+B$, unless the search class is restricted to local one-atom splits.

\paragraph{7. Guarded CA updates descend locally.}
Theorem~\ref{thm:refine-repair-descent:informal} states the algorithmic guarantee. A guarded repair does not increase the scoring-buffer objective because anchors and edges remain fixed and only a non-worsening local-state candidate is committed. An accepted refinement adds one concept and its actual accepted edge set only when its held-out penalized refinement gain beats the repair gain plus a margin and statistical safety term. Thus CA is not an unscored growth heuristic: it is a local refine-or-repair decision with an explicit gap comparison.

These statements also define the boundary of the claim. The scoring buffer is local and changes over time, so the descent result is not a global training theorem. The adaptive validation theorem requires fresh or otherwise unused held-out scoring samples; reused validation buffers need a reusable-holdout argument. The strict Bayes-risk refinement gap is a statement about the runtime information structure or a restricted hard-route prediction class; it is not a claim that adding CA edges increases information beyond a predictor that can already condition arbitrarily on the full hidden state. Approximate retrieval must satisfy candidate recall for Soft-CA edge gates to recover the formal Hard-CA route, and centroid repair may change the smooth Soft-CA read map even when the anchor-based hard information structure is fixed. Budgeting, merging, and eviction are systems approximations that may discard distinctions and must be evaluated separately.
