\section{Conclusion}
\label{sec:conclusion}

Runtime learning fails when it repairs errors that actually require new distinctions. In the refinable-partition model, this failure is a quotient error: a repair-only learner remains trapped inside the equivalence relation induced by its current state and suffers linear regret against a refined oracle whenever the refinement gap is positive. The sharper characterization is loss-relevant: a refinement matters when it changes the Bayes action, or under log loss when it changes the conditional outcome distribution. Distinction Graph Machines turn this separation into an explicit runtime learner with anchored concepts, calibrated local memories, distinction edges, and a read-before-observe update rule.

The theory explains why this separation matters. Contradictions inside an information cell cannot be solved by optimizing a predictor that still factors through the same information map. They require refinement. Once the local concept is adequate, its state can be repaired by sufficient-statistic, ridge, or projection-memory updates. DGM implements both operations in one state: concepts hold local memories, and distinction edges generate the information structure that gates the read path.

The resulting picture places gradient learning, local repair, and distinction refinement in one hierarchy. Backpropagation moves slow parameters; local memory repairs fast state; DGM adds the operation of changing the information structure on which those states are defined. The remaining empirical burden is substantial: the mechanism checks here should be extended to calibrated NLL, stronger online baselines, memory-latency budgets, and realistic frozen-model adaptation streams before making large-scale language-model claims.
