\section{Guarded Empirical Refinement}
\label{sec:main-guard}

Population value must be estimated before a runtime learner commits a new distinction.
For an atom $B$ and candidate split $g:B\to\{0,1\}$, the classification score used in the experiments is empirical log-loss improvement:
\[
\widehat V_{\log}(B,g)
=
H(\widehat p_B)
-
\sum_{b\in\{0,1\}}
\frac{|B_b|}{|B|}H(\widehat p_{B_b}),
\qquad
B_b=\{i\in B:g(x_i)=b\}.
\]
The analogous squared-loss score is $\operatorname{SSE}(B)-\operatorname{SSE}(B_0)-\operatorname{SSE}(B_1)$.

For a state $S$ and held-out scoring buffer $\calB_t^{\rm score}$ of size $m_t$, define
\[
\widehat R(S;\calB_t^{\rm score})
=
\frac1{m_t}\sum_{(\omega_i,Y_i)\in\calB_t^{\rm score}}
L(Y_i,f_S(\omega_i)),
\qquad
\widehat G_t(S')
=
\widehat R(S_t;\calB_t^{\rm score})
-
\widehat R(S';\calB_t^{\rm score}).
\]
Let $\calC_t^{\rm rep}$ be repair candidates and $\calC_t^{\rm ref}$ refinement candidates.
The best empirical repair gain is $\max_{S'\in\calC_t^{\rm rep}}\widehat G_t(S')$.
The best penalized refinement gain is $\max_{S'\in\calC_t^{\rm ref}}\{\widehat G_t(S')-\lambda c(S')\}$.
Candidate proposal may use a disjoint proposal buffer, but the scoring buffer must be held out from proposal and previous state-selection decisions for the concentration statement below.
This sample split is the statistical version of the paper's conceptual split: proposal signals can be broad and heuristic, but acceptance is a consequence-scored comparison between repair and refinement.
The identity state is included among repair candidates, so the best empirical repair gain is nonnegative and a harmful local update can be declined.

\begin{theorem}[Guarded empirical refinement]
\label{thm:main-guarded-refinement}
Assume losses lie in $[0,1]$.
At time $t$, let $\calC_t=\calC_t^{\rm rep}\cup\calC_t^{\rm ref}$ be predictable from the past and proposal buffer, and let $\calB_t^{\rm score}$ contain $m_t$ fresh held-out samples independent of candidate selection.
Define
\[
\epsilon_t
=
\sqrt{
\frac{
\log\bigl(4t^2(|\calC_t|+1)/\delta\bigr)
}{
2m_t
}
}.
\]
Then with probability at least $1-\delta$, uniformly over all $t$ and $S'\in\calC_t$,
\[
\left|\widehat G_t(S')-G_t(S')\right|
\le 2\epsilon_t.
\]
Consequently, if a refinement $S_f$ is accepted because
\[
\widehat G_t(S_f)-\lambda c(S_f)
>
\widehat G_t(S_r)+\tau+4\epsilon_t,
\]
where $S_r$ is the best empirical repair candidate, then
\[
G_t(S_f)-\lambda c(S_f)
>
G_t(S_r)+\tau.
\]
\end{theorem}

\begin{proof}[Proof sketch]
Condition on the past filtration; the old state and candidate states are fixed.
Hoeffding controls $|\widehat R(S)-R(S)|$ for the old state and all candidates, and a union bound over $|\calC_t|+1$ states and over $t$ with the $t^{-2}$ schedule gives the displayed high-probability event.
A gain is a difference of two risks, hence its deviation is at most $2\epsilon_t$.
Comparing a refinement gain to a repair gain subtracts two gain deviations, which is why the guard uses $4\epsilon_t$.
\end{proof}

The theorem is a finite-sample guard, not a global training theorem.
If the same validation buffer is reused adaptively, one needs reusable-holdout or adaptive-validation control; otherwise the rule should be read as an empirical heuristic.
The margin also clarifies the role of the threshold $\tau$.
The $4\epsilon_t$ term accounts for sampling uncertainty in the empirical comparison; $\tau$ is a modeling margin that can absorb unmodeled costs, numerical instability, or a desired minimum effect size.
In DGM, $c(S')$ charges concept and edge growth, while repair candidates keep the accepted distinction set fixed.
The result assumes a finite candidate class at the scoring time.
Appendix~\ref{sec:value} gives broader greedy-recovery and oracle-inequality statements, including a local-split bound whose atom count grows as $m_0+B$ rather than $m_0 2^B$.
