\section{Concept Graph Attention}
\label{sec:main-cga}

CGA is an attention layer whose memory is a fixed bank of runtime concepts
rather than a table of token keys and values.
For each sequence and layer, the state before block $b$ is
\[
\mathcal{M}_b =
\left\{
(a_j,s^k_{jh},s^v_{jh},n_{jh})_{h=1}^{H}
\right\}_{j=1}^{C}.
\]
Here $a_j$ is a fixed slot prior used to break symmetry, $s^k_{jh}$ is the key
sufficient statistic for slot $j$ and head $h$, $s^v_{jh}$ is the value
sufficient statistic, and $n_{jh}$ is a count or mass statistic.
All $C$ concept slots are available from the beginning of the sequence.
Each slot also has a fixed edge predicate
$e_j=(u_j,b_j,\sigma_j)$ derived from the random slot priors.
For the main implementation, $u_j$ is the separating direction between slot
$j$'s prior and its nearest neighboring prior, $b_j$ is the midpoint bias, and
$\sigma_j$ is the side containing slot $j$.
The slot addresses and memories used by the read are
\[
c_{jh}=\operatorname{normalize}\!\left(s^k_{jh}/\max(n_{jh},1)\right),
\qquad
m_{jh}=s^v_{jh}/\max(n_{jh},1).
\]
At initialization, $n_{jh}=n_0$, $s^k_{jh}=n_0a_{jh}$, and $s^v_{jh}=0$.
The concept budget $C$ is a model hyperparameter analogous to a recurrent-state
budget.
The value memory may use a concept bottleneck dimension $d_v^c$ smaller than
the ordinary attention head value dimension.
For address dimension $d_a$, value bottleneck $d_v^c$, and $H$ heads, the
persistent per-sequence state scales as
\[
S_{\rm CGA}
=
C\,H(d_a+d_v^c+1).
\]
The formula counts per-head key sums, value sums, and counts.
Slot priors and edge predicates are fixed layer buffers rather than
per-sequence runtime state.
The important scaling point is that $S_{\rm CGA}$ depends on $C$, not on the
context length.

\paragraph{What learns at inference time.}
During training, CGA learns slow parameters: projections, output maps, write
gates, and retention parameters.
During inference, these slow parameters are fixed.
What changes is the per-sequence concept memory $\mathcal{M}_b$.
After each block, the layer updates key statistics, value statistics, and count
statistics using the prefix it has just observed.
We call this \emph{test-time concept learning}: it is learning because later
predictions condition on a concept state estimated online from the test prefix,
but it is not test-time gradient descent on model weights.
This distinction is central to the method.
The hidden state is not a token cache, because writes aggregate information into
concept slots; it is not a learned inner-loop model, because the update is a
forward-only sufficient-statistic rule.

\paragraph{Dense concept read.}
Given normalized query $q_{th}$ and concept addresses $c_{jh}$, CGA reads by
continuous attention over edge-eligible concept slots.
Let
\[
\bar q_t=\operatorname{normalize}\!\left(\frac1H\sum_{h=1}^{H}q_{th}\right),
\qquad
E_t=\{j:\mathbf{1}[u_j^\top\bar q_t>b_j]=\sigma_j\}.
\]
If $E_t$ is empty, CGA falls back to all concept slots.
The read is
\[
\alpha_{thj}
=
\frac{\exp(\gamma q_{th}^{\top}c_{jh})}
{\sum_{r\in E_t}\exp(\gamma q_{th}^{\top}c_{rh})},
\qquad
z_{th}=\sum_{j\in E_t}\alpha_{thj}m_{jh}.
\]
The scale $\gamma$ is learned and initialized at the standard dot-product
attention scale.
The output projection maps the concatenated head readout
$(z_{t1},\ldots,z_{tH})$ back to $d_{\rm model}$.
Thus the read path is ordinary differentiable attention, but the keys and
values are concept memories rather than token memories.

\paragraph{Gated concept write.}
CGA uses a read-before-write causal protocol.
After producing $z_t$, the layer forms a write key, value, and gate:
\[
k_t=Kx_t,\qquad v_t=Vx_t,\qquad
g_t=\sigma(w_g^\top x_t+b_g).
\]
For MQAR we use a one-token delayed write key so that a key token can bind to
the following value token:
\[
k^{\rm write}_t=k_{t-1},\qquad v^{\rm write}_t=v_t.
\]
For general language modeling, the delay can be set to zero or treated as a
task-dependent write convention.
The write is routed to the most compatible concept slot:
\[
j_t
=
\arg\max_{j\le C}
\frac1H\sum_{h=1}^{H}(k^{\rm write}_{th})^\top c_{jh}
\]
The route $j_t$ is detached fast state.
The value write and gate remain differentiable through future concept reads.

\paragraph{Trainable retention.}
CGA maintains a trainable retention decay over old concept statistics.
For head $h$,
\[
\lambda_{th}
=
\exp(A_h)\,
\left(\delta_{\min}+(\delta_{\max}-\delta_{\min})
\sigma(u_h^\top x_t+b_h)\right)r_{th},
\qquad
\rho_{bh}=\exp\left(-\sum_{t\in b}\lambda_{th}\right),
\]
where $r_{th}=t^{-\alpha_h}$ when position-adaptive scaling is enabled and
$r_{th}=1$ otherwise.
The default parameterization of $A_h$ is PoST-style:
\[
A_h=A_0+\sum_{i<h}\operatorname{softplus}(\Delta_i),
\]
which gives ordered decay spectra across heads.
The implementation also exposes a free per-head $A_h$ parameterization for
ablation.
This decay plays the same role as the $A$ or retention gate in Mamba-style
state-space layers, but it acts on concept sufficient statistics rather than on
a fixed token-indexed recurrent state.

\paragraph{Block update.}
Within a block, CGA accumulates writes with scatter-add operations and commits
them at the block boundary:
\[
n_{j_t h} \leftarrow
\operatorname{sg}(\rho_{bh})n_{j_t h}
+\operatorname{sg}(g_t),
\]
\[
s^k_{j_t h} \leftarrow
\operatorname{sg}(\rho_{bh})s^k_{j_t h}
+\operatorname{sg}(g_t)k^{\rm write}_{th},
\]
\[
s^v_{j_t h} \leftarrow
\rho_{bh}s^v_{j_t h}
+g_t v^{\rm write}_{th}.
\]
Here $\operatorname{sg}$ denotes stop-gradient.
The key and count updates are detached routing state.
The value payload, write gate, and retention decay receive gradients through
future reads.
Because writes are sums over assigned observations and decay is multiplicative
per block, the update is scan-friendly: implementations can fuse dense concept
reads and use parallel scatter reductions for the write side.

\paragraph{Relation to ordinary attention.}
Let $K_{\rm tok},V_{\rm tok}$ be token-level keys and values.
Standard attention computes
\[
\operatorname{Attn}(X)_t
=
\sum_{s<t}
\operatorname{softmax}_s(q_t^\top K_{{\rm tok},s})V_{{\rm tok},s}.
\]
CGA instead computes
\[
\operatorname{CGA}(X)_t
=
\sum_{j\in E_t}
\operatorname{softmax}_j(q_t^\top c_{tj})m_{tj},
\]
where $c_{tj},m_{tj}$ are produced by the concept-bank update over the prefix.
Attention uses one memory entry per token event; CGA uses a bounded set of
concept slots that aggregate token evidence.
This is why CGA can be viewed as a memory-basis replacement for attention:
the read remains attention, but the state being attended to is concept-indexed
rather than token-indexed.

\paragraph{Learning and backpropagation boundary.}
CGA trains its slow parameters end-to-end:
embeddings, read/write projections, output projections, retention parameters,
and the language-model head receive gradient from the task loss.
The runtime concept update itself is not backpropagated through as a recurrent
state trajectory.
Routing decisions, key sums, counts, and slot priors are detached fast state.
The gated value content remains differentiable so that future concept reads can
train the value projection, write gate, and retention decay.
Thus CGA is neither a backprop-free language model nor a test-time optimizer
over model parameters.
It is an attention-compatible state layer whose fast memory update is a
forward-time associative learning operation while slow neural parameters are
trained by ordinary backpropagation.

\begin{algorithm}[t]
\caption{\textsc{ConceptGraphAttentionForward}}
\label{alg:main-cga}
\begin{algorithmic}[1]
\Statex Input: hidden states $x_{1:T}$, concept count $C$, block size $B$.
\State Initialize fixed concept bank with random priors $a_j$, edge predicates $e_j$, counts $n_0$, and zero values.
\For{blocks $[s,e)$}
    \State Project block queries, write keys, values, gates, and retention rates.
    \State Read edge-eligible concept slots with dense attention, falling back to all slots when needed.
    \State Emit block outputs through the output projection.
    \State Route each valid write to the most similar concept slot.
    \State Decay old sufficient statistics and scatter-add gated count, key, and value updates.
\EndFor
\State Return output projection of concept readouts.
\end{algorithmic}
\end{algorithm}
