\section{Related Work}
\label{sec:main-related}

\paragraph{Efficient sequence mixers and expressive state.}
Transformers use content-addressed attention over previous tokens \citep{vaswani2017attention}. Mamba, Mamba-2, RWKV, GDN, GLA, RetNet, and related models replace attention with efficient recurrent or linear state updates \citep{gu2023mamba,dao2024transformers,peng2023rwkv,yang2025gated,yang2024gated,sun2023retentive}. TTT-style models instead make the hidden state expressive by running a learned update rule at test time \citep{sun2025learning}. Our experiments place CA in this comparison class: CA is an attention-like causal mixer whose persistent state is a bounded concept memory rather than a token table.

\paragraph{Test-time learning layers.}
TTT layers frame sequence modeling as test-time learning: the hidden state is a small model, and the state update is a self-supervised training step on the current sequence \citep{sun2020test,sun2025learning}. CA adopts the same high-level goal, namely to let inference-time computation learn from the observed prefix, but changes the learning object. The hidden state in CA is not a linear model or MLP whose weights are updated; it is a concept-indexed value memory. The update is therefore forward-only at inference: read concepts, decay old values, and add gated continuous writes. This makes CA complementary to TTT. TTT explores expressive hidden states as trainable inner models; CA explores expressive hidden states as writable concept memories with attention-style readout.

\paragraph{Attention variants.}
Sparse attention, sliding-window attention, grouped-query attention, and compressed-memory attention mainly alter which token keys and values are read or how the token table is stored. CA is orthogonal: it changes the basis being attended to. The read still has the same softmax form, so optimized attention kernels and attention variants can in principle be adapted to concept slots instead of token positions.

\paragraph{Associative and external memory.}
Cache language models, $k$NN-LMs, retrieval-augmented systems, fast weights, online ridge or RLS memories, Memory Networks, Neural Turing Machines, and Differentiable Neural Computers all use forms of stateful retrieval or local memory update \citep{grave2017improving,khandelwal2020generalization,lewis2020retrieval,ba2016using,schlag2021linear,crammer2006online,weston2015memory,graves2014neural,graves2016hybrid}. CA shares the associative-memory goal but keeps the decoder-block interface of attention: each token reads from concept slots and writes back through a gated causal update.

\paragraph{Writable capacity.}
Mixture-of-experts, Switch Transformers, and Product-Key Memory route tokens or queries to trained capacity \citep{shazeer2017outrageously,fedus2022switch,lample2019large}. CA also weights writes across concept slots, but the writable capacity is runtime value state rather than a bank of trained expert parameters. The static concept count $C$ makes this capacity explicit and state-equalizable against recurrent mixers.
