\section{Single-Workstation Evaluation Grid and Diagnostics}
\label{app:additional-checks}

This appendix records the evaluation protocol that was moved out of the main
experiment section, fixes the intended single-workstation scale of the empirical
program, and records implementation checks for the non-oracle CIFAR-100
Image-HQS run. The run-tier grid is a design constraint; the Image-HQS audit
below summarizes the completed evidence package used in
Table~\ref{tab:cifar100-image-hqs}. The headline Image-HQS table was run as a
CUDA single-workstation experiment on an NVIDIA GeForce RTX 5090 Laptop GPU; the
method claim does not depend on Apple-specific hardware.

\subsection{Runtime contract and evaluation suite}

Every method receives the same frozen encoder or language model. The cost of
training or loading this shared model is reported but not charged to runtime
adaptation. At round $t$, the method receives $x_t$, computes the shared hidden
state $h_t=\phi_\theta(x_t)$, predicts $\hat y_t$, then observes the consequence
$y_t$ and may update only its declared runtime state:
\[
x_t \longrightarrow h_t \longrightarrow \hat y_t,
\qquad
y_t \longrightarrow \text{runtime update}.
\]
DGM updates by detached graph-memory repair or by guarded refinement. Test-time
training baselines update a head, prompt, adapter, normalization state, or other
declared adaptation parameters by inference-time gradient steps. The reported
adaptation overhead excludes the shared frozen forward pass and includes all
extra reads, writes, backward passes, activation storage, optimizer state,
synchronization, and memory-index maintenance.

\begin{table}[h]
\centering
\footnotesize
\begin{tabular}{>{\raggedright\arraybackslash}p{0.18\linewidth}
>{\raggedright\arraybackslash}p{0.28\linewidth}
>{\raggedright\arraybackslash}p{0.24\linewidth}
>{\raggedright\arraybackslash}p{0.18\linewidth}}
\toprule
Benchmark & What it tests & Single-workstation scale target & Primary success condition \\
\midrule
Hidden-context control & Frozen states intentionally collapse regimes that require different predictions. The oracle refined partition is known. & $10^6$--$10^7$ stream rounds, $K\in\{8,32,128\}$ regimes, label-noise grid, 5 seeds. & DGM approaches the refined oracle and beats repair-only regret slope. \\
Repair-only control & Same stream geometry but no hidden consequence conflict. Errors come from stale local statistics. & Matched stream length and concept budget. & DGM accepts few refinements and matches the best repair baseline. \\
Frozen vision streams & CIFAR-10/100-style frozen-encoder classification with class/domain bursts, corruptions, and replay-free label reveal. & Full CIFAR-10 and CIFAR-100 streams; optional larger local image subset if available. & Better prequential NLL or accuracy at matched adaptation time and memory. \\
Frozen LM streams & A frozen next-token model processes document continuations; the observed next token is the consequence. & $5$M--$50$M tokens with cached fp16 hidden states; char/word small model for smoke, GPT-style frozen model for max run. & Lower prequential NLL than cache and gradient-update adapters at lower backward cost. \\
Systems scaling & Synthetic hidden states plus labels isolate data-structure cost from model forward cost. & $N=10^3$--$10^6$ concepts, $d=128$--$2048$, top-$k=8$--$64$, edge degree $q=0$--$32$. & Report latency, memory, candidate recall, and refinement throughput. \\
\bottomrule
\end{tabular}
\caption{Evaluation suite. The controlled streams test the theorem-level
separation; the image and language streams test practical frozen-backbone
adaptation; the systems benchmark establishes the largest single-workstation
operating envelope.}
\label{tab:dgm-evaluation-suite}
\end{table}

\subsection{Run tiers}

\begin{table}[h]
\centering
\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.15\linewidth}
>{\raggedright\arraybackslash}p{0.26\linewidth}
>{\raggedright\arraybackslash}p{0.28\linewidth}
>{\raggedright\arraybackslash}p{0.20\linewidth}}
\toprule
Tier & Purpose & Scale & Use in paper \\
\midrule
Smoke & Catch implementation bugs and check predict-before-observe order. &
$10^4$ synthetic rounds, CIFAR-10 subset, $10^5$ language tokens, one seed. &
Never used as headline evidence. \\
Main & Default evidence tier for all tables. &
$10^6$ controlled rounds, full CIFAR-10/100 streams, $5$M--$10$M language tokens,
3--5 seeds. & Main paper tables and figures. \\
Max & Largest single-workstation run used for scaling claims. &
$10^7$ controlled rounds or synthetic states, $50$M cached-token hidden states
when storage permits, up to $10^6$ concepts in systems tests. &
Systems and scalability figure. \\
\bottomrule
\end{tabular}
\caption{Experiment tiers. Smoke runs are engineering checks; main and max runs
are the only tiers that should support empirical claims.}
\label{tab:single-workstation-run-tiers}
\end{table}

\subsection{Hardware and accounting}

The hardware report must include CPU, accelerator, accelerator memory or unified
memory, operating system, Python version, framework version, backend, dtype, and
whether hidden states are recomputed or cached. For every method, report both
total wall clock and the runtime-adaptation overhead after the shared frozen
forward pass. For gradient baselines, count backward passes, optimizer state,
and activation storage for the adaptation window. For DGM and cache baselines,
count concept or example memory, edge memory, index memory, and
validation-buffer memory.

The max run should stop before memory pressure or swapping dominates. The memory
cap should be declared before the run as a fixed fraction of available RAM or
accelerator memory. Exceeding the cap is a failed systems configuration rather
than a valid larger experiment.

\subsection{Controlled-stream grid}

The hidden-context control should sweep regime count, noise, and drift:
\[
K\in\{8,32,128\},\qquad
\epsilon_{\mathrm{label}}\in\{0,0.05,0.15\},\qquad
T\in\{10^6,10^7\}.
\]
The repair-only control uses the same marginal $x_t$ process and stream length
but removes the hidden label conflict. Required curves are cumulative
prequential regret, accepted concept count, accepted edge count, and held-out
split value. A convincing DGM run has a sublinear-looking regret curve relative
to the refined oracle after discovery, while repair-only baselines keep a
positive asymptotic slope on the hidden-context stream and do not on the
repair-only stream.

\subsection{Vision-stream grid}

The main vision run uses a frozen encoder and three observation orders:
random-balanced, class-burst, and corruption/domain-burst. CIFAR-10 is for fast
debugging; CIFAR-100 is the main mandatory dataset. If a local larger image
dataset is available, an ImageNet-100-style subset can be added as a max run,
but it should not replace CIFAR-100 because reviewers must be able to reproduce
the baseline tier easily.

The main table should include frozen, online head SGD/Adam, online ridge or RLS,
budgeted $k$NN/cache, repair-only DGM, edge-free DGM, and full DGM. Report
prequential NLL, prequential accuracy, post-adaptation test NLL and accuracy,
adaptation seconds per 10k examples, peak memory, and accepted concepts/edges.

\subsection{Non-oracle Image-HQS audit and ablations}
\label{app:image-hqs-audit}

The Image-HQS implementation was audited for label leakage. The online DGM
update in the prequential loop calls the memory with only the current feature
vector and revealed binary consequence. For the posterior variant, the repair
routine receives only the query mask, the binary label, and the soft-routing
weight. The CIFAR fine and coarse labels enter only after training, inside
purity, true-coordinate rank, oracle diagnostic, and representation-ceiling
computations. Thus the headline non-oracle result does not update
$\pi_{j,c}$ from the true coarse label.

The small ablation suite in Table~\ref{tab:image-hqs-ablation} uses
$n_{\mathrm{train}}=2000$, $n_{\mathrm{dev}}=800$, and $n_{\mathrm{test}}=800$.
It is not the headline 10-seed result, but it checks whether the gain comes from
the posterior local memory rather than from a trivial global statistic.

\begin{table}[h]
\centering
\small
\begin{tabular}{lcc}
\toprule
Variant & Test NLL $\downarrow$ & Test accuracy $\uparrow$ \\
\midrule
Categorical DGM & $0.6878$ & $0.5450$ \\
Single-bit image-only DGM & $0.6911$ & $0.5325$ \\
Global posterior memory & $0.6695$ & $0.5925$ \\
Image-route posterior DGM & $0.6627$ & $0.5850$ \\
Image-route posterior DGM, $\eta_{\mathrm{read}}=0$ & $0.6629$ & $0.5850$ \\
Image+mask-route posterior DGM & $0.6623$ & $0.5913$ \\
\bottomrule
\end{tabular}
\caption{Posterior-memory ablations for non-oracle CIFAR-100 Image-HQS. The
known-noise readout is not driving the result: fixing the posterior readout
noise to zero changes NLL by only $0.0002$ in this ablation.}
\label{tab:image-hqs-ablation}
\end{table}

The global posterior is stronger than frequency because it can learn mask
coordinate biases from the binary stream, but it remains worse than local
posterior DGM. The categorical and single-bit variants fail because learned
image concepts are not coarse-pure enough for one accepted mask bit to represent
the local Bayes action. The image+mask routing ablation is included as a
diagnostic only; the main result uses image-only routing because it most closely
matches the causal structure ``image chooses a coarse coordinate, mask queries
that coordinate.''

\subsection{Language-stream grid}

The language benchmark uses read-before-observe next-token prediction. The
small run may use a character or word model trained locally, but the main run
should use a frozen pretrained or fixed checkpoint model so that DGM is judged
as an inference-time layer rather than as extra offline training. The stream is
long contiguous text, not randomly shuffled tokens. The comparison set is:
frozen model, continuous cache or $k$NN-LM, online soft-prompt/head/adapter
updates, repair-only DGM-Embedding, and full DGM-Embedding.

The max run may precompute hidden states in fp16 and replay the adaptation
layer, provided an end-to-end online subset verifies identical predictions up to
floating-point tolerance. Report token NLL, perplexity, adaptation microseconds
per token, peak memory, backward passes per token, and concept/edge growth over
document position.

\subsection{Evidence boundary}

For the non-oracle CIFAR-100 Image-HQS result, the paper claims predictive
evidence under a shared frozen encoder: DGM posterior memory improves test NLL
and accuracy over the reported online linear baselines without label leakage.
It does not claim a full loss-latency-memory frontier against stronger
nonparametric and inference-time gradient baselines. That broader frontier is a
future-work requirement rather than a premise of
Table~\ref{tab:cifar100-image-hqs}.

\subsection{Required failure diagnostics}

The following checks are required before interpreting a positive result:
\begin{itemize}
    \item \textbf{Order check:} the observed label or token must not enter any
    prediction made for that same round.
    \item \textbf{Backbone equality:} all methods must use identical frozen
    hidden states for the same stream.
    \item \textbf{Budget equality:} cache, $k$NN, and DGM memory budgets must be
    matched in bytes, not just in number of stored items.
    \item \textbf{Candidate recall:} controlled streams should report how often
    the oracle-compatible concept is present in the sparse candidate set.
    \item \textbf{Growth sanity:} on repair-only controls, full DGM should not
    grow a large graph; otherwise the refinement guard is too weak.
    \item \textbf{Ablation consistency:} if edge-free DGM matches full DGM on the
    hidden-context control, the experiment is not testing graph distinctions.
\end{itemize}

These diagnostics are stricter than ordinary benchmark reporting because the
paper's claim is structural. A result that improves accuracy while failing the
order, budget, or ablation checks is not evidence for the refine-or-repair
claim.
