\section{Projection Learning Proofs}
\label{app:projection-proofs}

\begin{lemma}[Bregman three-point identity]
\label{lem:three-point:formal}
For all $a,b,u\in S$,
\begin{align*}
D_\psi(u,a)-D_\psi(u,b)-D_\psi(b,a)=\ip{\nabla\psi(a)-\nabla\psi(b)}{b-u}.
\end{align*}
\end{lemma}
\begin{proof}
Expand each term:
\begin{align*}
D_\psi(u,a)&=\psi(u)-\psi(a)-\ip{\nabla\psi(a)}{u-a},\\
D_\psi(u,b)&=\psi(u)-\psi(b)-\ip{\nabla\psi(b)}{u-b},\\
D_\psi(b,a)&=\psi(b)-\psi(a)-\ip{\nabla\psi(a)}{b-a}.
\end{align*}
Then
\begin{align*}
D_\psi(u,a)-D_\psi(u,b)-D_\psi(b,a)&=-\ip{\nabla\psi(a)}{u-a}+\ip{\nabla\psi(b)}{u-b}+\ip{\nabla\psi(a)}{b-a}\\
&=\ip{\nabla\psi(a)}{b-u}-\ip{\nabla\psi(b)}{b-u}\\
&=\ip{\nabla\psi(a)-\nabla\psi(b)}{b-u}.
\end{align*}
\end{proof}

\begin{theorem}[Projection characterization, formal version of Theorem~\ref{thm:projection-characterization:informal}]
\label{thm:projection-characterization:formal}
Let $S$ be closed and convex, let $\psi$ be differentiable and strictly convex, let $\calC\subseteq S$ be closed and convex, and let $s_0\in S$. For $p\in\calC$, the following are equivalent:
\begin{itemize}
\item {\bf Part 1.} $p=\argmin_{s\in\calC}D_\psi(s,s_0)$.
\item {\bf Part 2.} For all $u\in\calC$,
\begin{align*}
D_\psi(u,p)+D_\psi(p,s_0)\le D_\psi(u,s_0).
\end{align*}
\end{itemize}
\end{theorem}
\begin{proof}
Assume Part 1. The first-order optimality condition is
\begin{align*}
\ip{\nabla\psi(p)-\nabla\psi(s_0)}{u-p}\ge0\qquad \forall u\in\calC.
\end{align*}
Equivalently,
\begin{align*}
\ip{\nabla\psi(s_0)-\nabla\psi(p)}{p-u}\ge0\qquad \forall u\in\calC.
\end{align*}
By Lemma~\ref{lem:three-point:formal},
\begin{align*}
D_\psi(u,s_0)-D_\psi(u,p)-D_\psi(p,s_0)=\ip{\nabla\psi(s_0)-\nabla\psi(p)}{p-u}.
\end{align*}
Thus Part 2 follows.

Assume Part 2. The same identity implies
\begin{align*}
\ip{\nabla\psi(s_0)-\nabla\psi(p)}{p-u}\ge0\qquad \forall u\in\calC.
\end{align*}
Hence
\begin{align*}
\ip{\nabla\psi(p)-\nabla\psi(s_0)}{u-p}\ge0\qquad \forall u\in\calC.
\end{align*}
This is the variational inequality for minimizing $D_\psi(s,s_0)$ over $\calC$. Strict convexity gives uniqueness, so Part 1 follows.
\end{proof}

\begin{corollary}[Fejer monotonicity and bounded drift]
\label{cor:fejer:formal}
Let $s_{t+1}=\argmin_{s\in\calC_t}D_\psi(s,s_t)$. If $u^\star\in\cap_{t=1}^T\calC_t$, then
\begin{align*}
D_\psi(u^\star,s_{t+1})\le D_\psi(u^\star,s_t)
\end{align*}
and
\begin{align*}
\sum_{t=1}^T D_\psi(s_{t+1},s_t)\le D_\psi(u^\star,s_1).
\end{align*}
\end{corollary}
\begin{proof}
By Theorem~\ref{thm:projection-characterization:formal},
\begin{align*}
D_\psi(u^\star,s_{t+1})+D_\psi(s_{t+1},s_t)\le D_\psi(u^\star,s_t).
\end{align*}
Dropping the second term gives the first claim. Summing gives
\begin{align*}
\sum_{t=1}^T D_\psi(s_{t+1},s_t)&\le \sum_{t=1}^T\left(D_\psi(u^\star,s_t)-D_\psi(u^\star,s_{t+1})\right)\\
&=D_\psi(u^\star,s_1)-D_\psi(u^\star,s_{T+1})\\
&\le D_\psi(u^\star,s_1).
\end{align*}
\end{proof}

\begin{theorem}[Soft projection inequality]
\label{thm:soft-projection:formal}
Let $\ell_t:S\to\R$ be convex and let
\begin{align*}
s_{t+1}=\argmin_{s\in S}\{D_\psi(s,s_t)+\eta_t\ell_t(s)\}.
\end{align*}
Then for all $u\in S$,
\begin{align*}
\eta_t(\ell_t(s_{t+1})-\ell_t(u))\le D_\psi(u,s_t)-D_\psi(u,s_{t+1})-D_\psi(s_{t+1},s_t).
\end{align*}
\end{theorem}
\begin{proof}
For some $g_{t+1}\in\partial\ell_t(s_{t+1})$, optimality gives
\begin{align*}
\ip{\nabla\psi(s_{t+1})-\nabla\psi(s_t)+\eta_t g_{t+1}}{u-s_{t+1}}\ge0.
\end{align*}
Thus
\begin{align*}
\eta_t\ip{g_{t+1}}{s_{t+1}-u}\le \ip{\nabla\psi(s_t)-\nabla\psi(s_{t+1})}{s_{t+1}-u}.
\end{align*}
Convexity gives
\begin{align*}
\ell_t(s_{t+1})-\ell_t(u)\le \ip{g_{t+1}}{s_{t+1}-u}.
\end{align*}
Lemma~\ref{lem:three-point:formal} gives
\begin{align*}
\ip{\nabla\psi(s_t)-\nabla\psi(s_{t+1})}{s_{t+1}-u}=D_\psi(u,s_t)-D_\psi(u,s_{t+1})-D_\psi(s_{t+1},s_t).
\end{align*}
Combining proves the claim.
\end{proof}

\begin{theorem}[Soft projection as implicit mirror descent, formal version of Theorem~\ref{thm:implicit-md:informal}]
\label{thm:implicit-md:formal}
Let $\ell_t$ be differentiable and let
\begin{align*}
s_{t+1}=\argmin_s\{D_\psi(s,s_t)+\eta\ell_t(s)\}.
\end{align*}
If $s_{t+1}$ is an interior optimum, then
\begin{align*}
\nabla\psi(s_{t+1})=\nabla\psi(s_t)-\eta\nabla\ell_t(s_{t+1}).
\end{align*}
If $\psi(s)=\norm{s}_2^2/2$, then
\begin{align*}
s_{t+1}=s_t-\eta\nabla\ell_t(s_{t+1}).
\end{align*}
\end{theorem}
\begin{proof}
At an interior optimum,
\begin{align*}
0=\nabla_sD_\psi(s_{t+1},s_t)+\eta\nabla\ell_t(s_{t+1}).
\end{align*}
Since
\begin{align*}
\nabla_sD_\psi(s,s_t)=\nabla\psi(s)-\nabla\psi(s_t),
\end{align*}
we obtain
\begin{align*}
\nabla\psi(s_{t+1})-\nabla\psi(s_t)+\eta\nabla\ell_t(s_{t+1})=0.
\end{align*}
Rearranging proves the mirror form. If $\psi(s)=\norm{s}_2^2/2$, then $\nabla\psi(s)=s$.
\end{proof}

\begin{theorem}[Quadratic projection memory as implicit gradient descent, formal version of Theorem~\ref{thm:pml-implicit:informal}]
\label{thm:pml-implicit:formal}
Let $\ell(W)=\norm{Wk-v}^2/2$. The update
\begin{align*}
W_{t+1}=\argmin_W\left\{\frac{1}{2}\norm{W-W_t}_F^2+\eta\ell(W)\right\}
\end{align*}
is equivalent to
\begin{align*}
W_{t+1}=W_t-\eta\nabla\ell(W_{t+1})
\end{align*}
and has closed form
\begin{align*}
W_{t+1}=W_t+\frac{\eta}{1+\eta\norm{k}^2}(v-W_tk)k^\top.
\end{align*}
Moreover, if $e=v-W_tk$, then
\begin{align*}
v-W_{t+1}k=\frac{1}{1+\eta\norm{k}^2}e.
\end{align*}
\end{theorem}
\begin{proof}
The gradient is
\begin{align*}
\nabla\ell(W)=(Wk-v)k^\top.
\end{align*}
The optimality condition for the proximal objective is
\begin{align*}
0=W_{t+1}-W_t+\eta(W_{t+1}k-v)k^\top,
\end{align*}
which is the implicit gradient step. Let $e=v-W_tk$ and set $W_{t+1}=W_t+\alpha ek^\top$. Then
\begin{align*}
v-W_{t+1}k=e-\alpha e\norm{k}^2=(1-\alpha\norm{k}^2)e.
\end{align*}
The implicit step gives
\begin{align*}
\alpha ek^\top=\eta(1-\alpha\norm{k}^2)ek^\top.
\end{align*}
Thus
\begin{align*}
\alpha=\frac{\eta}{1+\eta\norm{k}^2}.
\end{align*}
Substituting $\alpha$ gives the update and residual contraction.
\end{proof}

\begin{theorem}[KL consistency projection]
\label{thm:kl-projection:formal}
Let $p_t$ be strictly positive on a finite space $\calX$. The solution of
\begin{align*}
\argmin_{p:\mathbb E_p[\phi]=c}D_{\mathrm{KL}}(p\|p_t)
\end{align*}
has the form
\begin{align*}
p_{t+1}(x)=\frac{p_t(x)\exp(\lambda^\top\phi(x))}{Z(\lambda)}.
\end{align*}
\end{theorem}
\begin{proof}
The Lagrangian is
\begin{align*}
\calL(p,\lambda,\nu)=\sum_x p(x)\log\frac{p(x)}{p_t(x)}+\lambda^\top\left(\sum_x p(x)\phi(x)-c\right)+\nu\left(\sum_xp(x)-1\right).
\end{align*}
Stationarity gives
\begin{align*}
\log\frac{p(x)}{p_t(x)}+1+\lambda^\top\phi(x)+\nu=0.
\end{align*}
Hence, after absorbing signs into $\lambda$ and normalizing,
\begin{align*}
p_{t+1}(x)=\frac{p_t(x)\exp(\lambda^\top\phi(x))}{Z(\lambda)}.
\end{align*}
\end{proof}

\begin{proposition}[Gradient descent as linearized projection]
\label{prop:gd-linearized:formal}
For $D_\psi(s,s_t)=\norm{s-s_t}_2^2/2$, replacing $\ell_t(s)$ in the soft projection objective by its first-order approximation at $s_t$ gives $s_{t+1}=s_t-\eta\nabla\ell_t(s_t)$.
\end{proposition}
\begin{proof}
The linearized objective is
\begin{align*}
\frac{1}{2}\norm{s-s_t}^2+\eta\ip{\nabla\ell_t(s_t)}{s-s_t}+\eta\ell_t(s_t).
\end{align*}
The stationarity condition is
\begin{align*}
s-s_t+\eta\nabla\ell_t(s_t)=0.
\end{align*}
Thus
\begin{align*}
s_{t+1}=s_t-\eta\nabla\ell_t(s_t).
\end{align*}
\end{proof}
